### 2024-05-30 Semantics-2024/paper.qmd
initial

---
title: Consolidating Wind Farm Data in Data Spaces for Training Predictive Maintenance ML

abstract: |
  The UNDERPIN project addresses the challenge of training machine learning (ML) models for predictive maintenance of wind farms by leveraging consolidated training data provided by multiple partners. This collaborative effort is facilitated through a European-wide Data Space for Manufacturing for dynamic asset management and predictive and prescriptive maintenance established as part of the project. Data spaces allow trusted participants to share data while maintaining data sovereignty. However, consolidating diverse data sets requires a clear semantics and mapping framework.
  To solve this, the project employs Semantic Web standards and technologies to model and process the data. Using OWL (Web Ontology Language), vocabularies and ontologies are developed to represent the data with clear, standardized semantics. Two key software components, GraphDB and PoolParty, are utilized to implement these standards and provide a semantic layer to efficiently process the consolidated data. The main benefit is to support ML training by presenting a consolidated and uniform training data set. Our approach aims to enhance the quality and reliability of predictive maintenance models for wind farms.
  A demonstration from the UNDERPIN project showcases the process of modeling and processing the data, illustrating the practical application of these semantic technologies. This example highlights the project's approach and the potential improvements in predictive maintenance through enhanced data integration and processing.

author:
  - name: Robert David
    email: 
    orcid: 
    affiliation: 
     name: The Semantic Web Company
     country: Austria
  - name: Axel Weißenfeld
    email: 
    orcid: 
    affiliation: 
      name: Austrian Institute of Technology
      country: Austria
  - name: Petar Ivanov
    email: 
    orcid: 
  - name: Nikola Tulechki
    email: 
    orcid: 
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
    affiliation: 
      name: Ontotext
      country: Bulgaria
acm-metadata:
  keywords:
    - data spaces
    - oil and gas
    - renewable energy
    - refineries
    - windfarms
    - time series
    - ontologies
    - semantic technologies
  final: true
  # acmart-options: sigconf, review
  # copyright-year: 2024
  # acm-year: 2024
  copyright: acmcopyright
  # doi: 
  # conference-acronym: 
  # conference-name: 
  # conference-date: 
  # conference-location: 
  # isbn:

format:
  acm-pdf:
    keep-tex: true 
bibliography: bibliography.bib

---

# Introduction
Consolidating wind farm data in data spaces for training predictive maintenance ML - Use Case Description

# Initial Situation
In the UNDERPIN project, we face the challenge of training ML for predictive maintenance of wind farms. This scenario benefits from large amounts of training data. Such data can be provided by different partners who collaborate on this ML approach. We support this scenario by implementing a data space for wind farm data. Data spaces enable data sharing between trusted participants while keeping data sovereignty. However, for consolidating different data sets from participants, we need to establish a clear semantics and mapping for the data.

# Approach and IT-Solution
To tackle this problem, we introduce Semantic Web standards and technologies to model and process this data. Vocabularies and ontologies, specifically expressed using OWL, can represent such data with a clear semantics based on standards. Furthermore, we introduce GraphDB and PoolParty as two software components which implement these standards to process the data.

The Wind Farm Ontology [@ikeresnaola-gonzalezWindFarmOntology2021] ...

# Business Value and Benefits of the Semantic Solution
When consolidating the data, we need to provide a method to automatically map and process it so this can be done efficiently on large data sets and/or regular updates in data spaces. The benefit provided should support the ML training in such a way that the consolidated data set is exposed as uniform training data set.

# Prospects and Recommendations
We provide a first prototype to implement our approach. Next steps are bringing it into the actual data space in practice and representing and integrating further data sets from different data space participants. Also, the ML should actually benefit regarding quality in practice, which still needs to be verified.

# Conclusion

# References

### 2024-05-30 Semantics-2024/paper.tex
initial

\documentclass[manuscript,screen]{acmart}

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother

%%
%% This is file `sample-manuscript.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `manuscript')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-manuscript.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

%% PANDOC PREAMBLE BEGINS

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
%% PANDOC PREAMBLE ENDS

\setlength{\parindent}{10pt}
\setlength{\parskip}{0pt}

\hypersetup{
  pdftitle={Consolidating Wind Farm Data in Data Spaces for Training Predictive Maintenance ML},
  pdfauthor={Robert David; Axel Weißenfeld; Petar Ivanov; Nikola Tulechki; Vladimir Alexiev},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={red},
  pdfcreator={LaTeX via pandoc, via quarto}}

%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{}
\acmYear{}
\acmDOI{}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[]{}{}{}
\acmPrice{}
\acmISBN{}

%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Consolidating Wind Farm Data in Data Spaces for Training
Predictive Maintenance ML}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

  \author{Robert David}

            \affiliation{%
                  \institution{The Semantic Web Company}
                                                  \country{Austria}
                      }
        \author{Axel Weißenfeld}

            \affiliation{%
                  \institution{Austrian Institute of Technology}
                                                  \country{Austria}
                      }
        \author{Petar Ivanov}

    \author{Nikola Tulechki}

    \author{Vladimir Alexiev}
  \orcid{0000-0001-7508-7428}
            \affiliation{%
                  \institution{Ontotext}
                                                  \country{Bulgaria}
                      }

%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}
%%  
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
The UNDERPIN project addresses the challenge of training machine
learning (ML) models for predictive maintenance of wind farms by
leveraging consolidated training data provided by multiple partners.
This collaborative effort is facilitated through a European-wide Data
Space for Manufacturing for dynamic asset management and predictive and
prescriptive maintenance established as part of the project. Data spaces
allow trusted participants to share data while maintaining data
sovereignty. However, consolidating diverse data sets requires a clear
semantics and mapping framework. To solve this, the project employs
Semantic Web standards and technologies to model and process the data.
Using OWL (Web Ontology Language), vocabularies and ontologies are
developed to represent the data with clear, standardized semantics. Two
key software components, GraphDB and PoolParty, are utilized to
implement these standards and provide a semantic layer to efficiently
process the consolidated data. The main benefit is to support ML
training by presenting a consolidated and uniform training data set. Our
approach aims to enhance the quality and reliability of predictive
maintenance models for wind farms. A demonstration from the UNDERPIN
project showcases the process of modeling and processing the data,
illustrating the practical application of these semantic technologies.
This example highlights the project's approach and the potential
improvements in predictive maintenance through enhanced data integration
and processing.    
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{data spaces, oil and gas, renewable
energy, refineries, windfarms, time series, ontologies, semantic
technologies}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\setlength{\parskip}{-0.1pt}

\section{Introduction}\label{introduction}

Consolidating wind farm data in data spaces for training predictive
maintenance ML - Use Case Description

\section{Initial Situation}\label{initial-situation}

In the UNDERPIN project, we face the challenge of training ML for
predictive maintenance of wind farms. This scenario benefits from large
amounts of training data. Such data can be provided by different
partners who collaborate on this ML approach. We support this scenario
by implementing a data space for wind farm data. Data spaces enable data
sharing between trusted participants while keeping data sovereignty.
However, for consolidating different data sets from participants, we
need to establish a clear semantics and mapping for the data.

\section{Approach and IT-Solution}\label{approach-and-it-solution}

To tackle this problem, we introduce Semantic Web standards and
technologies to model and process this data. Vocabularies and
ontologies, specifically expressed using OWL, can represent such data
with a clear semantics based on standards. Furthermore, we introduce
GraphDB and PoolParty as two software components which implement these
standards to process the data.

The Wind Farm Ontology \citep{ikeresnaola-gonzalezWindFarmOntology2021}
\ldots{}

\section{Business Value and Benefits of the Semantic
Solution}\label{business-value-and-benefits-of-the-semantic-solution}

When consolidating the data, we need to provide a method to
automatically map and process it so this can be done efficiently on
large data sets and/or regular updates in data spaces. The benefit
provided should support the ML training in such a way that the
consolidated data set is exposed as uniform training data set.

\section{Prospects and
Recommendations}\label{prospects-and-recommendations}

We provide a first prototype to implement our approach. Next steps are
bringing it into the actual data space in practice and representing and
integrating further data sets from different data space participants.
Also, the ML should actually benefit regarding quality in practice,
which still needs to be verified.

\section{Conclusion}\label{conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography.bib}

%% begin pandoc before-bib
%% end pandoc before-bib
%% begin pandoc biblio
%% end pandoc biblio
%% begin pandoc include-after
%% end pandoc include-after
%% begin pandoc after-body
%% end pandoc after-body

\end{document}
\endinput
%%
%% End of file `sample-manuscript.tex'.

### 2024-06-03 Semantics-2024/paper.qmd
add orcid in paper.qmd

@@
   - name: Petar Ivanov
-    email: 
-    orcid: 
+    email: petar.ivanov@ontotext.com
+    orcid: 0000-0001-8448-1005
   - name: Nikola Tulechki

### 2024-06-03 EBDVF-2024/paper.qmd
start EBDVF-2024/paper.qmd

--- 2024-05-30 Semantics-2024/paper.qmd
@@
 ---
-title: Consolidating Wind Farm Data in Data Spaces for Training Predictive Maintenance ML
+title: Raising the role of Vocabulary Hubs for Semantic data interoperability in Dataspaces
 
 abstract: |
-  The UNDERPIN project addresses the challenge of training machine learning (ML) models for predictive maintenance of wind farms by leveraging consolidated training data provided by multiple partners.
-This collaborative effort is facilitated through a European-wide Data Space for Manufacturing for dynamic asset management and predictive and prescriptive maintenance established as part of the project.
-Data spaces allow trusted participants to share data while maintaining data sovereignty.
-However, consolidating diverse data sets requires a clear semantics and mapping framework.
-  To solve this, the project employs Semantic Web standards and technologies to model and process the data.
-Using OWL (Web Ontology Language), vocabularies and ontologies are developed to represent the data with clear, standardized semantics.
-Two key software components, GraphDB and PoolParty, are utilized to implement these standards and provide a semantic layer to efficiently process the consolidated data.
-The main benefit is to support ML training by presenting a consolidated and uniform training data set.
-Our approach aims to enhance the quality and reliability of predictive maintenance models for wind farms.
-  A demonstration from the UNDERPIN project showcases the process of modeling and processing the data, illustrating the practical application of these semantic technologies.
-This example highlights the project's approach and the potential improvements in predictive maintenance through enhanced data integration and processing.
+  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
+  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
+  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
+  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated data.
+  The Vocabulary Hub service (part of the IDSA RAM) provides a repository for ontologies and vocabularies.
+  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
+  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
+  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
+  leveraging the partnership between the Semantic Web Company and Ontotext,
+  and is being developed within the frame of the Digital Europe project UNDERPIN,
+  with applications to refinery and wind farm data.
 
 author:
+  - name: Martin Kaltenböck
+    email: martin.kaltenboeck@semantic-web.com
   - name: Robert David
-    email: 
-    orcid: 
+    email: robert.david@semantic-web.com
+    orcid: 0000-0002-3244-5341
     affiliation: 
@@
      country: Austria
-  - name: Axel Weißenfeld
-    email: 
-    orcid: 
-    affiliation: 
-      name: Austrian Institute of Technology
-      country: Austria
-  - name: Petar Ivanov
-    email: 
-    orcid: 
-  - name: Nikola Tulechki
-    email: 
-    orcid: 
   - name: Vladimir Alexiev
@@
     orcid: 0000-0001-7508-7428
+  - name: Petar Ivanov
+    email: petar.ivanov@ontotext.com
+    orcid: 0000-0001-8448-1005
     affiliation: 
@@
     - data spaces
+    - semantic interoperability
+    - semantic technologies
+    - ontologies
+    - vocabulary hub
     - oil and gas
@@
     - windfarms
-    - time series
-    - ontologies
-    - semantic technologies
   final: true

### 2024-06-10 EBDVF-2024/paper.qmd
first draft of semantic layer

@@
 
+## IDS Vocabulary Hub
+IDS defines the architecture of data spaces in the reference architecture model IDS-RAM.
+One of the defined architectural components, the Vocabulary Hub, has the role to provide vocabularies to annotate and describe data assets and services.
+These vocabularies form a common language to enable semantic interoperability among the data space participants.
+The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of a standardized semantic descriptions.
+It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies and thereby provides semantic  interoperability in a standardized and machine-readable way.
+
+## GraphDB Graph Database
+tbd
+
+## PoolParty Semenatic Suite
+PoolParty Semantic Suite is a semantic middleware platform based on W3C standards, specifically the Semantic Web, which provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.
+
+At the core of PoolParty is the Thesaurus Manager, which can be used to create and maintain RDF-based vocabularies, including ontologies, schemas and different kinds of knowledge models.
+These vocabularies can be applied for various use cases, from bridging data silos in enterprises by implementing data consolidation via the vocabularies, to annotating documents with semantic annotations to implement semantic search and recommendations which provide insights into data.
+
+PoolParty is well suited to form the basis of a vocabulary management system within a data space, effectively implementing not only the Vocabulary Hub role, but furthermore providing additional services which improve the expressivity of common vocabularies and automate semantic interoperability by intelligent data analysis.
+
+## The GraphDB Poolparty Integration
+The integration of GraphDB and PoolParty leverages synergies of a high-performance graph database with a platform for knowledge graph management.
+
+Both GraphDb and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL.
+They therefore seamlessly integrate with each other, but also with other services and components in data spaces which support these machine-readable standards for semantic interoperability.
+
 The Wind Farm Ontology [@ikeresnaola-gonzalezWindFarmOntology2021] ...
@@
 The benefit provided should support the ML training in such a way that the consolidated data set is exposed as uniform training data set.
+
+## the Semantic Layer Approach
+The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in data spaces by provisioning common vocabularies.
+We aim to go one step further and envision a Semantic Layer for data spaces, which provide vocabulary-based services for advanced and automated semantic descriptions of metadata and data.
+The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.
+
+In the following, we discuss i) the services provided for the Vocabulary Hub role and ii) additional services as extensions to the Vocabulary Hub, which implement towards a Semantic Layer for data spaces.
+
+Vocabulary Hub Services:
+* RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
+* scalable storage of vocabularies: cluster implementation for higher throughput and a fault tolerant architecture.
+* knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs.
+Vocabularies can be created and edited based on standards like SKOS and OWL, but also existing vocabularies can be reused and combined to fit specific needs of data spaces.
+* SPARQL service for semantic lookups: dereference of URIs describing data asset to retrieve a semantic description of the entity from the graph database.
+
+Semantic Layer Services:
+* we support both unstructured and structured data assets
++ unstructured data: semantic annotation services for documents based on concept tagging.
+PoolParty’s KG-based NLP  primarily uses SKOS taxonomies to identify concepts in documents, but can be extended with OWL ontology elements (classes, properties) for increased expressivity.
++ structured data: we support methods for structured data which can be decided based on the actual content in place, ranging from OWL ontologies for OBDA, linking SKOS concepts and
+- with OBDA for data spaces we support an ontology-based method which is well suited for data retrieval while keeping the original source in place.
+Ontologies for OBDA are intended to be governed via the Semantic Layer and can consolidate data assets of various formats using a precise mapping.
+Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
+- R2RML mapping is intended for data transformation processes within a data space, where the original source is used as a basis to create an RDF representation.
+- SKOS concepts can also be used for structured data in several ways.
+Concepts can be directly added to RDF data via semantic links (properties from an ontology), they can be used to annotate specific literals of an RDF graph or they can be used to associate documents with (specific parts of) an RDF graph.
+* Inference services: we provide metadata expansions services based on interlinked vocabularies in combination with ontologies.
+This inference approach can be used with the semantic tagging for unstructured data as wells as the structured data approaches.
++ Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inferences. The query can be configured to work for various scenarios.
+* Vocabulary crossovers are a data modeling approach using interlinks from an ontology to relate
++ different vocabularies from similar or the same domains, which cover different areas or aspects of the domain.
+By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand the descriptions by traversing the interlinked graph.
++ different versions of the same vocabulary to automatically translate between those versions.
+* Metadata inference based on interlinked vocabularies is similar to crossovers, but intended to represent different levels of abstraction, so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa.
+For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go more into detail.
 

### 2024-06-11 EBDVF-2024/paper.qmd
Added introduction and current situation (projects and use cases).

@@
 # Introduction
-Consolidating wind farm data in data spaces for training predictive maintenance ML - Use Case Description
+The european data economy heavily depends on the availability of data and the technological foundation to make use of it.
+Different industries and communities have a mutual benefit of sharing data between each other and thereby support their business goals.
+Examples are the well-known mobility data space, but also use cases like energy communities for running simulations and machine learning based systems, which heavily rely on high amount of high qualitative training data.
+To cope with the challenge of data sharing, data space approaches like IDS were introduced to not only solve the technical aspects of data sharing, but also to establish methods for data sovereignty so that data space participants can determine themselves how and when others can make use of the shared data.
+For making use of shared data, interoperability is a crucial factor.
+While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability.
+IDS supports semantic interoperability by leveraging Semantic Web standards and architectural components, like the Vocabulary Hub, which stores semantic vocabularies for metadata and data descriptions and can dereference URIs to provide semantic details of data assets.
+In this paper, we present our proposal for a technological solution of an IDS Vocabulary Hub build on the integration of the GraphDB graph database and PoolParty Semantic Suite.
+We not only cover the core Vocabulary Hub functionalities and responsibilities in a data space, but also extend the idea towards a Semantic Layer for data spaces, which provides even more services to easily and semi-automatically manage semantic interoperability.
 
-# Initial Situation
-In the UNDERPIN project, we face the challenge of training ML for predictive maintenance of wind farms.
-This scenario benefits from large amounts of training data.
-Such data can be provided by different partners who collaborate on this ML approach.
-We support this scenario by implementing a data space for wind farm data.
-Data spaces enable data sharing between trusted participants while keeping data sovereignty.
-However, for consolidating different data sets from participants, we need to establish a clear semantics and mapping for the data.
+# Projects and Use Cases
+In the expanding data economy landscape supported by data spaces, we are working in several projects which address the practical needs and were we experience the importance of semantic interoperability to fulfill various use cases.
+Here we present 3 examples of from the DataBri-X and UNDERPIN projects, which cover various application domains.
+
+## The Energy Data Space
+In this use case, we run simulations for energy communities, which predict the behavior of the energy grid.
+To provide precise predictions, there is the need of high amounts of example data.
+This data is of numeric type and different tabular data sources need to be semantically consolidated.
+
+## The Legal Data Space
+Analyzing documents in the legal domain is a highly challenging task, because of the complexity of legal information.
+We provide recommender services for insights into corpora of legal documents, which identify the meaning via a legal knowledge graph, NLP analysis and semantic annotations.
+
+## The Wind Farm Data Space
+Operating wind farms involves high costs for maintaining the wind turbines.
+Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business critical.
+For establishing predictive maintenance, we need high amounts of high qualitative training data, which can be achieved by consolidation different wind farm measurement sources.
+Again, we have the case where we need semantic interoperability for numeric tabular data.
+
+These are just 3 examples to illustrate the heterogeneity of data in the context of different use cases.
+We aim to develop solutions which not only fulfill the data space needs, but also improve the situation regarding semantic interoperability by providing services which can be easily leveraged to implement such use cases.
 

### 2024-06-12 EBDVF-2024/paper.qmd
Added project uris and line breaks

@@
 Different industries and communities have a mutual benefit of sharing data between each other and thereby support their business goals.
-Examples are the well-known mobility data space, but also use cases like energy communities for running simulations and machine learning based systems, which heavily rely on high amount of high qualitative training data.
-To cope with the challenge of data sharing, data space approaches like IDS were introduced to not only solve the technical aspects of data sharing, but also to establish methods for data sovereignty so that data space participants can determine themselves how and when others can make use of the shared data.
+Examples are the well-known mobility data space, but also use cases like energy communities for running simulations
+and machine learning based systems, which heavily rely on high amount of high qualitative training data.
+To cope with the challenge of data sharing, data space approaches like IDS were introduced to not only solve the technical aspects
+of data sharing, but also to establish methods for data sovereignty so that data space participants can determine themselves
+how and when others can make use of the shared data.
 For making use of shared data, interoperability is a crucial factor.
 While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability.
-IDS supports semantic interoperability by leveraging Semantic Web standards and architectural components, like the Vocabulary Hub, which stores semantic vocabularies for metadata and data descriptions and can dereference URIs to provide semantic details of data assets.
-In this paper, we present our proposal for a technological solution of an IDS Vocabulary Hub build on the integration of the GraphDB graph database and PoolParty Semantic Suite.
-We not only cover the core Vocabulary Hub functionalities and responsibilities in a data space, but also extend the idea towards a Semantic Layer for data spaces, which provides even more services to easily and semi-automatically manage semantic interoperability.
+IDS supports semantic interoperability by leveraging Semantic Web standards and architectural components, like the Vocabulary Hub,
+which stores semantic vocabularies for metadata and data descriptions and can dereference URIs to provide semantic details of data assets.
+In this paper, we present our proposal for a technological solution of an IDS Vocabulary Hub build on the integration of
+the GraphDB graph database and PoolParty Semantic Suite.
+We not only cover the core Vocabulary Hub functionalities and responsibilities in a data space, but also extend the idea
+towards a Semantic Layer for data spaces, which provides even more services to easily and semi-automatically manage semantic interoperability.
 
 # Projects and Use Cases
-In the expanding data economy landscape supported by data spaces, we are working in several projects which address the practical needs and were we experience the importance of semantic interoperability to fulfill various use cases.
-Here we present 3 examples of from the DataBri-X and UNDERPIN projects, which cover various application domains.
+In the expanding data economy landscape supported by data spaces, we are working in several projects which address the practical needs
+and were we experience the importance of semantic interoperability to fulfill various use cases.
+Here we present 3 examples of from the [DataBri-X](https://databri-x.eu/) and [UNDERPIN](https://underpinproject.eu/) projects,
+which cover various application domains.
 
@@
 Analyzing documents in the legal domain is a highly challenging task, because of the complexity of legal information.
-We provide recommender services for insights into corpora of legal documents, which identify the meaning via a legal knowledge graph, NLP analysis and semantic annotations.
+We provide recommender services for insights into corpora of legal documents, which identify the meaning via a legal knowledge graph,
+NLP analysis and semantic annotations.
 
@@
 Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business critical.
-For establishing predictive maintenance, we need high amounts of high qualitative training data, which can be achieved by consolidation different wind farm measurement sources.
+For establishing predictive maintenance, we need high amounts of high qualitative training data, which can be achieved
+by consolidation different wind farm measurement sources.
 Again, we have the case where we need semantic interoperability for numeric tabular data.
@@
 These are just 3 examples to illustrate the heterogeneity of data in the context of different use cases.
-We aim to develop solutions which not only fulfill the data space needs, but also improve the situation regarding semantic interoperability by providing services which can be easily leveraged to implement such use cases.
+We aim to develop solutions which not only fulfill the data space needs, but also improve the situation
+regarding semantic interoperability by providing services which can be easily leveraged to implement such use cases.
 
@@
 IDS defines the architecture of data spaces in the reference architecture model IDS-RAM.
-One of the defined architectural components, the Vocabulary Hub, has the role to provide vocabularies to annotate and describe data assets and services.
+One of the defined architectural components, the Vocabulary Hub, has the role to provide vocabularies to annotate and describe
+data assets and services.
 These vocabularies form a common language to enable semantic interoperability among the data space participants.
-The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of a standardized semantic descriptions.
-It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies and thereby provides semantic  interoperability in a standardized and machine-readable way.
+The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form
+of a standardized semantic descriptions.
+It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies and thereby provides
+semantic interoperability in a standardized and machine-readable way.
 
@@
 ## PoolParty Semenatic Suite
-PoolParty Semantic Suite is a semantic middleware platform based on W3C standards, specifically the Semantic Web, which provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.
+PoolParty Semantic Suite is a semantic middleware platform based on W3C standards, specifically the Semantic Web,
+which provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP,
+semantic search and recommender systems.
 
-At the core of PoolParty is the Thesaurus Manager, which can be used to create and maintain RDF-based vocabularies, including ontologies, schemas and different kinds of knowledge models.
-These vocabularies can be applied for various use cases, from bridging data silos in enterprises by implementing data consolidation via the vocabularies, to annotating documents with semantic annotations to implement semantic search and recommendations which provide insights into data.
+At the core of PoolParty is the Thesaurus Manager, which can be used to create and maintain RDF-based vocabularies,
+including ontologies, schemas and different kinds of knowledge models.
+These vocabularies can be applied for various use cases, from bridging data silos in enterprises by implementing data consolidation
+via the vocabularies, to annotating documents with semantic annotations to implement semantic search and recommendations
+which provide insights into data.
 
-PoolParty is well suited to form the basis of a vocabulary management system within a data space, effectively implementing not only the Vocabulary Hub role, but furthermore providing additional services which improve the expressivity of common vocabularies and automate semantic interoperability by intelligent data analysis.
+PoolParty is well suited to form the basis of a vocabulary management system within a data space,
+effectively implementing not only the Vocabulary Hub role, but furthermore providing additional services which improve
+the expressivity of common vocabularies and automate semantic interoperability by intelligent data analysis.
 
@@
 
-Both GraphDb and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL.
-They therefore seamlessly integrate with each other, but also with other services and components in data spaces which support these machine-readable standards for semantic interoperability.
-
-The Wind Farm Ontology [@ikeresnaola-gonzalezWindFarmOntology2021] ...
+Both GraphDB and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL.
+They therefore seamlessly integrate with each other, but also with other services and components in data spaces which support
+these machine-readable standards for semantic interoperability.
 
 # Business Value and Benefits of the Semantic Solution
-When consolidating the data, we need to provide a method to automatically map and process it so this can be done efficiently on large data sets and/or regular updates in data spaces.
+When consolidating the data, we need to provide a method to automatically map and process it so this can be done efficiently
+on large data sets and/or regular updates in data spaces.
 The benefit provided should support the ML training in such a way that the consolidated data set is exposed as uniform training data set.
@@
 ## the Semantic Layer Approach
-The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in data spaces by provisioning common vocabularies.
-We aim to go one step further and envision a Semantic Layer for data spaces, which provide vocabulary-based services for advanced and automated semantic descriptions of metadata and data.
+The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability
+in data spaces by provisioning common vocabularies.
+We aim to go one step further and envision a Semantic Layer for data spaces, which provide vocabulary-based services
+for advanced and automated semantic descriptions of metadata and data.
 The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.
 
-In the following, we discuss i) the services provided for the Vocabulary Hub role and ii) additional services as extensions to the Vocabulary Hub, which implement towards a Semantic Layer for data spaces.
+In the following, we discuss i) the services provided for the Vocabulary Hub role and ii) additional services as extensions to the Vocabulary Hub,
+which implement towards a Semantic Layer for data spaces.
 

### 2024-06-12 EBDVF-2024/paper.qmd
edits

--- 2024-06-10 EBDVF-2024/paper.qmd
@@
 ---
-title: Raising the role of Vocabulary Hubs for Semantic data interoperability in Dataspaces
+title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
 
@@
 author:
-  - name: Martin Kaltenböck
-    email: martin.kaltenboeck@semantic-web.com
   - name: Robert David
@@
   keywords:
-    - data spaces
+    - dataspaces
     - semantic interoperability
@@
 # Introduction
-Consolidating wind farm data in data spaces for training predictive maintenance ML - Use Case Description
+The European data economy heavily depends on the availability of data and the technological foundation to make use of it. 
+Different industries and communities can benefit mutually by sharing data with each other, thereby supporting their digitization business goals.
+Machine-learning systems heavily rely on high volumes of high-quality training data. 
+Examples are the well-known mobility dataspace and use cases like energy communities for running simulations.
+To cope with the challenge of data sharing, Dataspace approaches like the IDS RAM were introduced to not only solve the technical aspects of data sharing, 
+but also to establish methods for data sovereignty so that dataspace participants can determine themselves how and when others can make use of their data.
 
-# Initial Situation
-In the UNDERPIN project, we face the challenge of training ML for predictive maintenance of wind farms.
-This scenario benefits from large amounts of training data.
-Such data can be provided by different partners who collaborate on this ML approach.
-We support this scenario by implementing a data space for wind farm data.
-Data spaces enable data sharing between trusted participants while keeping data sovereignty.
-However, for consolidating different data sets from participants, we need to establish a clear semantics and mapping for the data.
+For making use of shared data, interoperability is a crucial factor. 
+While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability.
+IDS RAM is firmly based on semantic metadata for all aspects and actors of a dataspace.
+Furthermore, it can leverage semantic interoperability approaches through the Vocabulary Hub
+that stores ontologies and semantic thesauri for data descriptions and can dereference URIs to provide semantic details of data assets.
+
+In this paper, we present our proposal for a technological solution of an IDS Vocabulary Hub 
+built on the integration of the GraphDB semantic graph database and PoolParty Semantic Suite. 
+We not only cover the core Vocabulary Hub functionalities and responsibilities in a dataspace, 
+but also extend the idea towards a Semantic Layer for dataspaces, 
+which provides even more services to easily and semi-automatically manage semantic interoperability.
+Thus we raise the role of the Vocabulary Hub to not only store semantic assets, 
+but also to bind semantic descriptions dynamically to incoming data, 
+therefore enabling richer discovery and semantic data integration between datasets.
+
+# Projects and Use Cases
+In the expanding data economy landscape supported by dataspaces, 
+we are working on several projects which address practical data sharing needs and were we experience the importance of semantic interoperability to fulfill various use cases. 
+Here we present 3 examples from the [DataBri-X](https://databri-x.eu/) and [UNDERPIN](https://underpinproject.eu/) projects that cover various application domains.
+
+## Energy Dataspace
+In this use case, we run simulations for energy communities, which predict the behavior of the energy grid. 
+To provide precise predictions, there is the need of high amounts of example data. 
+This data is of numeric type and different tabular data sources need to be semantically consolidated.
+
+## Legal Dataspace
+Analyzing documents in the legal domain is a highly challenging task because of the complexity of legal information. 
+We provide recommender services for insights into corpora of legal documents
+that identify the meaning via a legal knowledge graph, NLP analysis and semantic annotations.
+
+## Manufacturing/Maintenance Dataspace
+Operating refineries and wind farms involves high maintenance costs. 
+Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business-critical. 
+For establishing predictive maintenance, we need high amounts of high quality training data
+that can be obtained by data consolidation across different refinery and wind farm measurement sources.
+
+The UNDERPIN project addresses this problem by creating a dataspace for refineries, wind farms, and their equipment suppliers.
+Extensive time series data is exchanged, but their format and meaning is not standardized.
+Describing the meaning of these time series facilitates semantic interoperability between datasets from different providers.
+This description includes measured variable, quantity kind and unit of measure,
+which sensor took the measurements, what it is attached to, the detailed model and connectivity of that industrial equipment.
+
+## The Problem
+"Standard" dataspace solutions provide access to data, but do not typically address the data integration problem.
+If a consumer wants to consume datasets from different providers in an integrated way, 
+they need to harmonize the data to some common model, convert it, store it in a database, 
+and potentially implement entity linking (correlation of different records that are about the same real-world thing) and data fusion.
+
+Above we gave just 3 examples to illustrate the heterogeneity of data in the context of different use cases. 
+We aim to develop solutions that not only fulfill the dataspace needs, 
+but also improve the situation regarding semantic interoperability by providing services that can be easily leveraged to implement such use cases.
 
 # Approach and IT-Solution
-To tackle this problem, we introduce Semantic Web standards and technologies to model and process this data.
-Vocabularies and ontologies, specifically expressed using OWL, can represent such data with a clear semantics based on standards.
-Furthermore, we introduce GraphDB and PoolParty as two software components which implement these standards to process the data.
+To tackle this problem, we introduce Semantic Web standards and technologies to model and process data. 
+Vocabularies and ontologies (expressed using RDFS, OWL and SKOS) can represent data with a clear semantics based on standards. 
+We leverage GraphDB and PoolParty as two software components that implement these standards to process the data.
 
 ## IDS Vocabulary Hub
-IDS defines the architecture of data spaces in the reference architecture model IDS-RAM.
+IDSA defines the architecture of dataspaces in the reference architecture model IDS RAM. 
 One of the defined architectural components, the Vocabulary Hub, has the role to provide vocabularies to annotate and describe data assets and services. 
-These vocabularies form a common language to enable semantic interoperability among the data space participants.
-The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of a standardized semantic descriptions.
-It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies and thereby provides semantic  interoperability in a standardized and machine-readable way.
+These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
+The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions. 
+It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies 
+and thereby provides semantic interoperability in a standardized and machine-readable way.
 
 ## GraphDB Graph Database
-tbd
+TODO
 
-## PoolParty Semenatic Suite
+## PoolParty Semantic Suite
 PoolParty Semantic Suite is a semantic middleware platform based on W3C standards, specifically the Semantic Web, which provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.
@@
 
-PoolParty is well suited to form the basis of a vocabulary management system within a data space, effectively implementing not only the Vocabulary Hub role, but furthermore providing additional services which improve the expressivity of common vocabularies and automate semantic interoperability by intelligent data analysis.
+PoolParty is well suited to form the basis of a vocabulary management system within a dataspace, 
+effectively implementing not only the Vocabulary Hub role, 
+but also providing additional services that improve the expressiveness of common vocabularies and automate semantic interoperability.
 
@@
 Both GraphDb and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL.
-They therefore seamlessly integrate with each other, but also with other services and components in data spaces which support these machine-readable standards for semantic interoperability.
+They therefore seamlessly integrate with each other, but also with other services and components in dataspaces which support these machine-readable standards for semantic interoperability.
 
-The Wind Farm Ontology [@ikeresnaola-gonzalezWindFarmOntology2021] ...
+# A Semantic Solution and its Benefits
+When harmonizing data, we need to provide a method to automatically map and process it,
+so this can be done efficiently on large datasets and/or regular updates in dataspaces. 
+The benefit provided should support the ML training in such a way that the consolidated dataset is exposed as uniform training dataset.
 
-# Business Value and Benefits of the Semantic Solution
-When consolidating the data, we need to provide a method to automatically map and process it so this can be done efficiently on large data sets and/or regular updates in data spaces.
-The benefit provided should support the ML training in such a way that the consolidated data set is exposed as uniform training data set.
-
-## the Semantic Layer Approach
-The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in data spaces by provisioning common vocabularies.
-We aim to go one step further and envision a Semantic Layer for data spaces, which provide vocabulary-based services for advanced and automated semantic descriptions of metadata and data.
+## Semantic Layer Approach
+The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
+We aim to go one step further and envision a Semantic Layer for dataspaces, 
+which provides vocabulary-based services for advanced and automated semantic descriptions of metadata and data. 
 The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.
 
-In the following, we discuss i) the services provided for the Vocabulary Hub role and ii) additional services as extensions to the Vocabulary Hub, which implement towards a Semantic Layer for data spaces.
+In the following, we discuss 
+i) the services provided for the Vocabulary Hub component and 
+ii) additional services as Vocabulary Hub extensions that implement a Semantic Layer for dataspaces.
 
 Vocabulary Hub Services:
-* RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
-* scalable storage of vocabularies: cluster implementation for higher throughput and a fault tolerant architecture.
-* knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs.
-Vocabularies can be created and edited based on standards like SKOS and OWL, but also existing vocabularies can be reused and combined to fit specific needs of data spaces.
-* SPARQL service for semantic lookups: dereference of URIs describing data asset to retrieve a semantic description of the entity from the graph database.
+- RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
+- Scalable vocabulary storage: cluster implementation for higher throughput and a fault tolerant architecture.
+- Knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs. 
+  Vocabularies can be created and edited based on standards like RDFS, OWL and SKOS, and existing vocabularies can be reused and combined to fit specific dataspace needs.
+- SPARQL service for semantic lookups: dereference URIs describing data assets to retrieve a semantic description of the entity from the graph database.
 
-Semantic Layer Services:
-* we support both unstructured and structured data assets
-+ unstructured data: semantic annotation services for documents based on concept tagging.
-PoolParty’s KG-based NLP  primarily uses SKOS taxonomies to identify concepts in documents, but can be extended with OWL ontology elements (classes, properties) for increased expressivity.
-+ structured data: we support methods for structured data which can be decided based on the actual content in place, ranging from OWL ontologies for OBDA, linking SKOS concepts and
-- with OBDA for data spaces we support an ontology-based method which is well suited for data retrieval while keeping the original source in place.
-Ontologies for OBDA are intended to be governed via the Semantic Layer and can consolidate data assets of various formats using a precise mapping.
+Semantic Layer Services support both unstructured and structured data assets:
+- Unstructured data: semantic annotation services for documents based on concept tagging. 
+  PoolParty’s KG-based NLP primarily uses SKOS taxonomies to identify concepts in documents, 
+  but can be extended with OWL ontology elements (classes, properties) for increased expressiveness.
+- Structured data: we support structured data harmonization methods, including semantic mapping and linking SKOS concepts
+  - Ontology-Based Data Access (OBDA) is used for virtualization of relational data, i.e. accessing relational data as semantic data.
+    It is well suited for semantic querying of large amounts of data, while keeping the original relational data in place (NoETL approach). 
+    Ontologies are leveraged to construct the OBDA mapping; 
+    they are governed via the Semantic Layer and can harmonize data assets of various formats using a precise mapping. 
     Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
-- R2RML mapping is intended for data transformation processes within a data space, where the original source is used as a basis to create an RDF representation.
-- SKOS concepts can also be used for structured data in several ways.
-Concepts can be directly added to RDF data via semantic links (properties from an ontology), they can be used to annotate specific literals of an RDF graph or they can be used to associate documents with (specific parts of) an RDF graph.
-* Inference services: we provide metadata expansions services based on interlinked vocabularies in combination with ontologies.
-This inference approach can be used with the semantic tagging for unstructured data as wells as the structured data approaches.
-+ Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inferences. The query can be configured to work for various scenarios.
-* Vocabulary crossovers are a data modeling approach using interlinks from an ontology to relate
-+ different vocabularies from similar or the same domains, which cover different areas or aspects of the domain.
-By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand the descriptions by traversing the interlinked graph.
-+ different versions of the same vocabulary to automatically translate between those versions.
-* Metadata inference based on interlinked vocabularies is similar to crossovers, but intended to represent different levels of abstraction, so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa.
-For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go more into detail.
+  - RDB to RDF Mapping Language (R2RML) [@souripriyadasR2RMLRDBRDF2012]
+    and RDF Mapping Language (RML) [@RMLIntroduction2024]
+    are intended for data transformation processes within a dataspace, 
+    where the original source is used as a basis to create an RDF representation (ETL approach).
+  - GraphDB supports both OBDA and R2RML through its integration of ONTOP [@bagosiOntopFrameworkOntology2014].
+    The same mappings can be used for NoETL virtual data access, or for ETL (to materialize the semantic data).
+  - CSV on the Web (CSVW) [@swirrlCSVWStandards2024] is a set of W3C standards 
+    that allow detailed semantic description of tabular data using a JSON "manifest" file.
+  - SKOS concepts are used for structured data in several ways. 
+    Concepts can be directly added to RDF data via semantic links (properties from an ontology), 
+    they can be used to annotate specific literals of an RDF graph,
+    or they can be used to associate documents with (specific parts of) an RDF graph.
+- Inference services: we provide metadata expansion services based on interlinked vocabularies in combination with ontologies. 
+  This inference approach can be used with semantic tagging for unstructured data, as well as the structured data approaches.
+  - Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inference. 
+    The query can be configured to work for various scenarios.
+- Vocabulary crosswalks are a data modeling approach using interlinks to relate:
+  - Different vocabularies from similar or the same domain that cover different areas or aspects of the domain. 
+    By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand queries and annotations by traversing the interlinked graph.
+  - Different versions of the same vocabulary to automatically translate between those versions.
+- Metadata inference based on interlinked vocabularies is similar to crosswalks, 
+  but intended to represent different levels of abstraction, 
+  so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa. 
+  For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go into more detail.
 
-# Prospects and Recommendations
-We provide a first prototype to implement our approach.
-Next steps are bringing it into the actual data space in practice and representing and integrating further data sets from different data space participants.
+## Architecture and Implementation
+
+# Conclusion and Future Work
+We presented an approach for dynamically binding dataspace datasets
+We have implemented a first prototype of our approach.
+
+The next steps include:
+- Bringing it into an actual dataspace in practice, representing and integrating datasets from different dataspace participants.
+- Implementing use cases for richer discoverability and harmonized querying.
 Also, the ML should actually benefit regarding quality in practice, which still needs to be verified.
 
-# Conclusion
+# Acknowledgements
+This work is partially supported by the Digital Europe programme project [UNDERPIN](https://underpinproject.eu/) (grant agreement 101123179)
 

### 2024-06-12 EBDVF-2024/paper.qmd
minor

@@
 # Conclusion and Future Work
-We presented an approach for dynamically binding dataspace datasets
+We presented an approach for dynamically binding datasets to semantic descriptions in a dataspace's Vocabulary Hub,
+therefore facilitating data harmonization and easier data consumption.
 We have implemented a first prototype of our approach.

### 2024-06-12 EBDVF-2024/paper.qmd
add architecture description

@@
 
-## Architecture and Implementation
+## Software Architecture
+
+The following figure shows the architecture of our approach:
+
+![](img/architecture.png)
+
+- A Data Provider sends data through the Connector, 
+  and at the same time provides basic metadata (DCAT)
+- Selected data can be stored in Database(s) managed by the dataspace,
+  for easier access and querying.
+  - For large-volume time series data, we recommend to use InfluxDB, PostgreSQL, Google BigQuery, etc.
+  - Of course, access control is very important here.
+    We'll leverage GraphDB's [Fine-grained Access Control](https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html), 
+    but access control in the time series database is also important
+  - Depending on Provider preferences and agreements, 
+    other data may stay at the source, 
+    and be delivered to the Consumer only dynamically through the Connector
+- The Vocabulary Hub stores relevant semantic assets: ontologies and thesauri.
+  These are added based on relevant use cases and datasets, following a dataspace governance process.
+- The Vocabulary Hub also stores mapping assets, such as OBDA or R2RML mappings and CSVW manifests.
+- PoolParty Semantic Integrator is used to edit, view and manage the semantic assets,
+  and to bind mapping assets to incoming data.
+- For virtual access to relational data, ONTOP is used to translate SPARQL to SQL dynamically
+  and to convert returned data to SPARQL result format.
+- The Data Consumer benefits from:
+  - More powerful dataset discovery by using richer semantic queries, eg
+    - "Give me all time series related to compressor C123" (equipment instance)
+    - "Give me all time series related to temperature" (measured quantity)
+    - "Give me all time series of bearings" (equipment kind or machine part)
+  - Harmonized querying of datasets from multiple providers.
 

### 2024-06-12 EBDVF-2024/paper.qmd
Added details for use cases and conclusion

@@
 we are working on several projects which address practical data sharing needs and were we experience the importance of semantic interoperability to fulfill various use cases. 
-Here we present 3 examples from the [DataBri-X](https://databri-x.eu/) and [UNDERPIN](https://underpinproject.eu/) projects that cover various application domains.
-
-## Energy Dataspace
-In this use case, we run simulations for energy communities, which predict the behavior of the energy grid. 
-To provide precise predictions, there is the need of high amounts of example data. 
-This data is of numeric type and different tabular data sources need to be semantically consolidated.
-
-## Legal Dataspace
-Analyzing documents in the legal domain is a highly challenging task because of the complexity of legal information. 
-We provide recommender services for insights into corpora of legal documents
-that identify the meaning via a legal knowledge graph, NLP analysis and semantic annotations.
+our main focus to develop our solution is the [UNDERPIN](https://underpinproject.eu/) project with use cases of refineries and wind farms,
+but also other project, like [DataBri-X](https://databri-x.eu/) provide use cases which were considered.
 
@@
 which sensor took the measurements, what it is attached to, the detailed model and connectivity of that industrial equipment.
+
+### The Refinery Dataspace
+Refineries can be optimized regarding costs and efficiency by considering not only individual components,
+but the production chain as a connected system.
+In this use case, we aim to improve the maintenance process as well as the decision making for preventive maintenance
+so as to minimize the downtime and the impact on the production capabilities.
+
+### The Wind Farm Dataspace
+Operating wind farms involves high costs for maintaining the wind turbines.
+Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business critical.
+For establishing predictive maintenance, we need high amounts of high qualitative training data,
+which can be achieved by consolidation different wind farm measurement sources.
+This use case provides different kinds of tabular data, where consolidation needs semantic harmonization of data.
+
+## Energy and Legal Dataspace
+Beyond the UNDERPIN project as the main use case provider, we also consider use cases from other projects.
+We present 2 examples from the DataBri-X project, which work on different content types of data and have different usage scenarios.
+We considered these as well when designing our solution to make it applicable to a broader range of use cases.
+
+### The Energy Dataspace
+In this use case, we run simulations for energy communities, which predict the behavior of the energy grid. 
+To provide precise predictions, there is the need of high amounts of example data. 
+This data is of numeric type and different tabular data sources need to be semantically consolidated.
+
+### The Legal Dataspace
+Analyzing documents in the legal domain is a highly challenging task because of the complexity of legal information. 
+We provide recommender services for insights into corpora of legal documents
+that identify the meaning via a legal knowledge graph, NLP analysis and semantic annotations.
 
@@
 therefore facilitating data harmonization and easier data consumption.
-We have implemented a first prototype of our approach.
+We have implemented a first prototype of our approach, which builds on the integration of GraphDB and PoolParty.
 
@@
 - Bringing it into an actual dataspace in practice, representing and integrating datasets from different dataspace participants.
-- Implementing use cases for richer discoverability and harmonized querying.
+- Implementing use cases for richer discoverability, harmonized querying and support for different content types of structured and unstructured data.
 Also, the ML should actually benefit regarding quality in practice, which still needs to be verified.

### 2024-06-12 EBDVF-2024/paper.qmd
Added to next steps

@@
 We have implemented a first prototype of our approach, which builds on the integration of GraphDB and PoolParty.
+As next steps, we will bring this solution into use cases and into broader discussion to gain insight and develop it further.
 
 The next steps include:
-- Bringing it into an actual dataspace in practice, representing and integrating datasets from different dataspace participants.
+- Bringing it into actual dataspaces in practice, representing and integrating datasets from different dataspace participants.
 - Implementing use cases for richer discoverability, harmonized querying and support for different content types of structured and unstructured data.
-Also, the ML should actually benefit regarding quality in practice, which still needs to be verified.
+- Explore how ML can benefit regarding quality in practice when providing consolidated and cleaned data via dataspaces.
+- Discuss how we can extend the IDS-RAM with services to improve the support for semantic interoperability provided by our solution.
 

### 2024-06-12 EBDVF-2024/paper.qmd
Added to problem and solution

@@
 ## The Problem
-"Standard" dataspace solutions provide access to data, but do not typically address the data integration problem.
+"Standard" dataspace solutions provide access to data and metadata descriptions, but they do not typically address the data integration problem.
 If a consumer wants to consume datasets from different providers in an integrated way, 
@@
 
-Above we gave just 3 examples to illustrate the heterogeneity of data in the context of different use cases. 
+Above we gave just 4 examples to illustrate the heterogeneity of data in the context of different use cases.
 We aim to develop solutions that not only fulfill the dataspace needs, 
@@
 ## IDS Vocabulary Hub
-IDSA defines the architecture of dataspaces in the reference architecture model IDS RAM. 
+In this paper, we focus on the IDSA approach to dataspaces.
+IDSA defines the architecture of dataspaces in the reference architecture model IDS-RAM (current version 4).
 One of the defined architectural components, the Vocabulary Hub, has the role to provide vocabularies to annotate and describe data assets and services.

### 2024-06-14 EBDVF-2024/paper.qmd
Update paper.qmd

@@
   Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
-  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated data.
+  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
   The Vocabulary Hub service (part of the IDSA RAM) provides a repository for ontologies and vocabularies.
@@
 In the expanding data economy landscape supported by dataspaces, 
-we are working on several projects which address practical data sharing needs and were we experience the importance of semantic interoperability to fulfill various use cases. 
-our main focus to develop our solution is the [UNDERPIN](https://underpinproject.eu/) project with use cases of refineries and wind farms,
-but also other project, like [DataBri-X](https://databri-x.eu/) provide use cases which were considered.
+we are working on several projects which address practical data sharing needs and where we experience the importance of semantic interoperability to fulfill various use cases. 
+Our solution is developed with main focus on the [UNDERPIN](https://underpinproject.eu/) project with use cases of refineries and wind farms, 
+but other projects, such as [DataBri-X](https://databri-x.eu/) provide use cases which were considered.
 
@@
 Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business-critical. 
-For establishing predictive maintenance, we need high amounts of high quality training data
+For establishing predictive maintenance, we need large amounts of high quality training data 
 that can be obtained by data consolidation across different refinery and wind farm measurement sources.
 
-The UNDERPIN project addresses this problem by creating a dataspace for refineries, wind farms, and their equipment suppliers.
+The UNDERPIN project addresses this problem by creating a dataspace for European manufacturers in the refinery and renewable energy domains and various SMEs across their value-chain ecosystem, such as equipment suppliers.
 Extensive time series data is exchanged, but their format and meaning is not standardized. 
@@
 Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business critical. 
-For establishing predictive maintenance, we need high amounts of high qualitative training data,
-which can be achieved by consolidation different wind farm measurement sources.
+For establishing predictive maintenance, we need large amounts of high qualitative training data, 
+which can be achieved by consolidating different wind farm measurement sources. 
 This use case provides different kinds of tabular data, where consolidation needs semantic harmonization of data.
@@
 IDSA defines the architecture of dataspaces in the reference architecture model IDS-RAM (current version 4). 
-One of the defined architectural components, the Vocabulary Hub, has the role to provide vocabularies to annotate and describe data assets and services.
+One of the defined architectural components, the Vocabulary Hub, has the role of providing vocabularies to annotate and describe data assets and services.
 These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
@@
 
-Both GraphDb and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL.
+Both GraphDB and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL.
 They therefore seamlessly integrate with each other, but also with other services and components in dataspaces which support these machine-readable standards for semantic interoperability.
@@
 so this can be done efficiently on large datasets and/or regular updates in dataspaces. 
-The benefit provided should support the ML training in such a way that the consolidated dataset is exposed as uniform training dataset.
+The benefit provided should support the ML training in such a way that the consolidated dataset is exposed as a uniform training dataset.
 
@@
     We'll leverage GraphDB's [Fine-grained Access Control](https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html), 
-    but access control in the time series database is also important
+    but access control in the time series database is also important.
   - Depending on Provider preferences and agreements, 
     other data may stay at the source, 
-    and be delivered to the Consumer only dynamically through the Connector
+    and be delivered to the Consumer only dynamically through the Connector.
 - The Vocabulary Hub stores relevant semantic assets: ontologies and thesauri. 
@@
 - The Data Consumer benefits from:
-  - More powerful dataset discovery by using richer semantic queries, eg
+  - More powerful dataset discovery by using richer semantic queries, e.g.
     - "Give me all time series related to compressor C123" (equipment instance)

### 2024-06-14 EBDVF-2024/paper.tex
make paper.pdf

--- 2024-06-14 EBDVF-2024/paper.qmd
@@
----
-title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
+\documentclass[manuscript,screen]{acmart}
 
-abstract: |
-  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
-  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
-  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
-  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
-  The Vocabulary Hub service (part of the IDSA RAM) provides a repository for ontologies and vocabularies.
-  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
-  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
-  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
-  leveraging the partnership between the Semantic Web Company and Ontotext,
-  and is being developed within the frame of the Digital Europe project UNDERPIN,
-  with applications to refinery and wind farm data.
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
 
-author:
-  - name: Robert David
-    email: robert.david@semantic-web.com
-    orcid: 0000-0002-3244-5341
-    affiliation: 
-     name: The Semantic Web Company
-     country: Austria
-  - name: Vladimir Alexiev
-    email: vladimir.alexiev@ontotext.com
-    orcid: 0000-0001-7508-7428
-  - name: Petar Ivanov
-    email: petar.ivanov@ontotext.com
-    orcid: 0000-0001-8448-1005
-    affiliation: 
-      name: Ontotext
-      country: Bulgaria
-acm-metadata:
-  keywords:
-    - dataspaces
-    - semantic interoperability
-    - semantic technologies
-    - ontologies
-    - vocabulary hub
-    - oil and gas
-    - renewable energy
-    - refineries
-    - windfarms
-  final: true
-  # acmart-options: sigconf, review
-  # copyright-year: 2024
-  # acm-year: 2024
-  copyright: acmcopyright
-  # doi: 
-  # conference-acronym: 
-  # conference-name: 
-  # conference-date: 
-  # conference-location: 
-  # isbn:
+%%
+%% This is file `sample-manuscript.tex',
+%% generated with the docstrip utility.
+%%
+%% The original source files were:
+%%
+%% samples.dtx  (with options: `manuscript')
+%% 
+%% IMPORTANT NOTICE:
+%% 
+%% For the copyright see the source file.
+%% 
+%% Any modified versions of this file must be renamed
+%% with new filenames distinct from sample-manuscript.tex.
+%% 
+%% For distribution of the original source see the terms
+%% for copying and modification in the file samples.dtx.
+%% 
+%% This generated file may be distributed as long as the
+%% original source files, as listed above, are part of the
+%% same distribution. (The sources need not necessarily be
+%% in the same archive or directory.)
+%%
+%%
+%% Commands for TeXCount
+%TC:macro \cite [option:text,text]
+%TC:macro \citep [option:text,text]
+%TC:macro \citet [option:text,text]
+%TC:envir table 0 1
+%TC:envir table* 0 1
+%TC:envir tabular [ignore] word
+%TC:envir displaymath 0 word
+%TC:envir math 0 word
+%TC:envir comment 0 0
+%%
+%%
+%% The first command in your LaTeX source must be the \documentclass command.
 
-format:
-  acm-pdf:
-    keep-tex: true 
-bibliography: bibliography.bib
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 
----
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 
-# Introduction
-The European data economy heavily depends on the availability of data and the technological foundation to make use of it. 
-Different industries and communities can benefit mutually by sharing data with each other, thereby supporting their digitization business goals. 
-Machine-learning systems heavily rely on high volumes of high-quality training data. 
-Examples are the well-known mobility dataspace and use cases like energy communities for running simulations. 
-To cope with the challenge of data sharing, Dataspace approaches like the IDS RAM were introduced to not only solve the technical aspects of data sharing, 
-but also to establish methods for data sovereignty so that dataspace participants can determine themselves how and when others can make use of their data.
+%% PANDOC PREAMBLE BEGINS
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\definecolor{mypink}{RGB}{219, 48, 122}
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+%% PANDOC PREAMBLE ENDS
+
+\setlength{\parindent}{10pt}
+\setlength{\parskip}{0pt}
+
+\hypersetup{
+  pdftitle={Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces},
+  pdfauthor={Robert David; Vladimir Alexiev; Petar Ivanov},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={red},
+  pdfcreator={LaTeX via pandoc, via quarto}}
+
+%% \BibTeX command to typeset BibTeX logo in the docs
+\AtBeginDocument{%
+  \providecommand\BibTeX{{%
+    Bib\TeX}}}
+
+%% Rights management information.  This information is sent to you
+%% when you complete the rights form.  These commands have SAMPLE
+%% values in them; it is your responsibility as an author to replace
+%% the commands and values with those provided to you when you
+%% complete the rights form.
+\setcopyright{acmcopyright}
+\copyrightyear{}
+\acmYear{}
+\acmDOI{}
+
+%% These commands are for a PROCEEDINGS abstract or paper.
+\acmConference[]{}{}{}
+\acmPrice{}
+\acmISBN{}
+
+%% Submission ID.
+%% Use this when submitting an article to a sponsored event. You'll
+%% receive a unique submission ID from the organizers
+%% of the event, and this ID should be used as the parameter to this command.
+%%\acmSubmissionID{123-A56-BU3}
+
+%%
+%% For managing citations, it is recommended to use bibliography
+%% files in BibTeX format.
+%%
+%% You can then either use BibTeX with the ACM-Reference-Format style,
+%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
+%% support for advanced citation of software artefact from the
+%% biblatex-software package, also separately available on CTAN.
+%%
+%% Look at the sample-*-biblatex.tex files for templates showcasing
+%% the biblatex styles.
+%%
+
+%%
+%% The majority of ACM publications use numbered citations and
+%% references.  The command \citestyle{authoryear} switches to the
+%% "author year" style.
+%%
+%% If you are preparing content for an event
+%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
+%% citations and references.
+%% Uncommenting
+%% the next command will enable that style.
+%%\citestyle{acmauthoryear}
+
+%% end of the preamble, start of the body of the document source.
+\begin{document}
+
+%%
+%% The "title" command has an optional parameter,
+%% allowing the author to define a "short title" to be used in page headers.
+\title{Raising the Role of Vocabulary Hubs for Semantic Data
+Interoperability in Dataspaces}
+
+%%
+%% The "author" command and its associated commands are used to define
+%% the authors and their affiliations.
+%% Of note is the shared affiliation of the first two authors, and the
+%% "authornote" and "authornotemark" commands
+%% used to denote shared contribution to the research.
+
+  \author{Robert David}
+  \orcid{0000-0002-3244-5341}
+            \affiliation{%
+                  \institution{The Semantic Web Company}
+                                                  \country{Austria}
+                      }
+        \author{Vladimir Alexiev}
+  \orcid{0000-0001-7508-7428}
+    \author{Petar Ivanov}
+  \orcid{0000-0001-8448-1005}
+            \affiliation{%
+                  \institution{Ontotext}
+                                                  \country{Bulgaria}
+                      }
+
+%% By default, the full list of authors will be used in the page
+%% headers. Often, this list is too long, and will overlap
+%% other information printed in the page headers. This command allows
+%% the author to define a more concise list
+%% of authors' names for this purpose.
+%\renewcommand{\shortauthors}{Trovato et al.}
+%%  
+%% The abstract is a short summary of the work to be presented in the
+%% article.
+\begin{abstract}
+Dataspaces are an important enabler for industrial sharing data (either
+commercially licensed or private). Europe is investing heavily into
+sectoral dataspaces, federation and orchestration platforms like SIMPL,
+Eclipse DSC, GXFS, etc. Still, dataspaces enable shared data access, but
+do not solve the data interoperability problem. For that, the consumer
+would like to see the data from different providers in a harmonized and
+semantically integrated form. The Vocabulary Hub service (part of the
+IDSA RAM) provides a repository for ontologies and vocabularies. We
+describe an approach of raising the role of the vocabulary hub to also
+allow richer metadata description (e.g.~the meaning of every column in a
+tabular dataset), and binding semantic descriptions to ingested
+datasets, thus providing on-the-fly data semantization and easing data
+querying. This is achieved through the integration of two commercial
+semantic products (PoolParty and GraphDB), leveraging the partnership
+between the Semantic Web Company and Ontotext, and is being developed
+within the frame of the Digital Europe project UNDERPIN, with
+applications to refinery and wind farm data.    
+\end{abstract}
+
+%%
+%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
+%% Please copy and paste the code instead of the example below.
+%%
+
+%%
+%% Keywords. The author(s) should pick words that accurately describe
+%% the work being presented. Separate the keywords with commas.
+\keywords{dataspaces, semantic interoperability, semantic
+technologies, ontologies, vocabulary hub, oil and gas, renewable
+energy, refineries, windfarms}
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
+\maketitle
+
+\setlength{\parskip}{-0.1pt}
+
+\section{Introduction}\label{introduction}
+
+The European data economy heavily depends on the availability of data
+and the technological foundation to make use of it. Different industries
+and communities can benefit mutually by sharing data with each other,
+thereby supporting their digitization business goals. Machine-learning
+systems heavily rely on high volumes of high-quality training data.
+Examples are the well-known mobility dataspace and use cases like energy
+communities for running simulations. To cope with the challenge of data
+sharing, Dataspace approaches like the IDS RAM were introduced to not
+only solve the technical aspects of data sharing, but also to establish
+methods for data sovereignty so that dataspace participants can
+determine themselves how and when others can make use of their data.
 
 For making use of shared data, interoperability is a crucial factor.
-While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability. 
-IDS RAM is firmly based on semantic metadata for all aspects and actors of a dataspace. 
-Furthermore, it can leverage semantic interoperability approaches through the Vocabulary Hub 
-that stores ontologies and semantic thesauri for data descriptions and can dereference URIs to provide semantic details of data assets.
+While syntactic data exchange is well covered by clearly defined data
+formats, there is still the open challenge of semantic interoperability.
+IDS RAM is firmly based on semantic metadata for all aspects and actors
+of a dataspace. Furthermore, it can leverage semantic interoperability
+approaches through the Vocabulary Hub that stores ontologies and
+semantic thesauri for data descriptions and can dereference URIs to
+provide semantic details of data assets.
 
-In this paper, we present our proposal for a technological solution of an IDS Vocabulary Hub 
-built on the integration of the GraphDB semantic graph database and PoolParty Semantic Suite. 
-We not only cover the core Vocabulary Hub functionalities and responsibilities in a dataspace, 
-but also extend the idea towards a Semantic Layer for dataspaces, 
-which provides even more services to easily and semi-automatically manage semantic interoperability. 
-Thus we raise the role of the Vocabulary Hub to not only store semantic assets, 
-but also to bind semantic descriptions dynamically to incoming data, 
-therefore enabling richer discovery and semantic data integration between datasets.
+In this paper, we present our proposal for a technological solution of
+an IDS Vocabulary Hub built on the integration of the GraphDB semantic
+graph database and PoolParty Semantic Suite. We not only cover the core
+Vocabulary Hub functionalities and responsibilities in a dataspace, but
+also extend the idea towards a Semantic Layer for dataspaces, which
+provides even more services to easily and semi-automatically manage
+semantic interoperability. Thus we raise the role of the Vocabulary Hub
+to not only store semantic assets, but also to bind semantic
+descriptions dynamically to incoming data, therefore enabling richer
+discovery and semantic data integration between datasets.
 
-# Projects and Use Cases
-In the expanding data economy landscape supported by dataspaces, 
-we are working on several projects which address practical data sharing needs and where we experience the importance of semantic interoperability to fulfill various use cases. 
-Our solution is developed with main focus on the [UNDERPIN](https://underpinproject.eu/) project with use cases of refineries and wind farms, 
-but other projects, such as [DataBri-X](https://databri-x.eu/) provide use cases which were considered.
+\section{Projects and Use Cases}\label{projects-and-use-cases}
 
-## Manufacturing/Maintenance Dataspace
+In the expanding data economy landscape supported by dataspaces, we are
+working on several projects which address practical data sharing needs
+and where we experience the importance of semantic interoperability to
+fulfill various use cases. Our solution is developed with main focus on
+the \href{https://underpinproject.eu/}{UNDERPIN} project with use cases
+of refineries and wind farms, but other projects, such as
+\href{https://databri-x.eu/}{DataBri-X} provide use cases which were
+considered.
+
+\subsection{Manufacturing/Maintenance
+Dataspace}\label{manufacturingmaintenance-dataspace}
+
 Operating refineries and wind farms involves high maintenance costs.
-Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business-critical. 
-For establishing predictive maintenance, we need large amounts of high quality training data 
-that can be obtained by data consolidation across different refinery and wind farm measurement sources.
+Predicting potential failures and optimizing the maintenance operations
+to minimize the costs can be business-critical. For establishing
+predictive maintenance, we need large amounts of high quality training
+data that can be obtained by data consolidation across different
+refinery and wind farm measurement sources.
 
-The UNDERPIN project addresses this problem by creating a dataspace for European manufacturers in the refinery and renewable energy domains and various SMEs across their value-chain ecosystem, such as equipment suppliers.
-Extensive time series data is exchanged, but their format and meaning is not standardized. 
-Describing the meaning of these time series facilitates semantic interoperability between datasets from different providers. 
-This description includes measured variable, quantity kind and unit of measure, 
-which sensor took the measurements, what it is attached to, the detailed model and connectivity of that industrial equipment.
+The UNDERPIN project addresses this problem by creating a dataspace for
+European manufacturers in the refinery and renewable energy domains and
+various SMEs across their value-chain ecosystem, such as equipment
+suppliers. Extensive time series data is exchanged, but their format and
+meaning is not standardized. Describing the meaning of these time series
+facilitates semantic interoperability between datasets from different
+providers. This description includes measured variable, quantity kind
+and unit of measure, which sensor took the measurements, what it is
+attached to, the detailed model and connectivity of that industrial
+equipment.
 
-### The Refinery Dataspace
-Refineries can be optimized regarding costs and efficiency by considering not only individual components, 
-but the production chain as a connected system. 
-In this use case, we aim to improve the maintenance process as well as the decision making for preventive maintenance 
-so as to minimize the downtime and the impact on the production capabilities.
+\subsubsection{The Refinery Dataspace}\label{the-refinery-dataspace}
 
-### The Wind Farm Dataspace
-Operating wind farms involves high costs for maintaining the wind turbines. 
-Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business critical. 
-For establishing predictive maintenance, we need large amounts of high qualitative training data, 
-which can be achieved by consolidating different wind farm measurement sources. 
-This use case provides different kinds of tabular data, where consolidation needs semantic harmonization of data.
+Refineries can be optimized regarding costs and efficiency by
+considering not only individual components, but the production chain as
+a connected system. In this use case, we aim to improve the maintenance
+process as well as the decision making for preventive maintenance so as
+to minimize the downtime and the impact on the production capabilities.
 
-## Energy and Legal Dataspace
-Beyond the UNDERPIN project as the main use case provider, we also consider use cases from other projects. 
-We present 2 examples from the DataBri-X project, which work on different content types of data and have different usage scenarios. 
-We considered these as well when designing our solution to make it applicable to a broader range of use cases.
+\subsubsection{The Wind Farm Dataspace}\label{the-wind-farm-dataspace}
 
-### The Energy Dataspace
-In this use case, we run simulations for energy communities, which predict the behavior of the energy grid. 
-To provide precise predictions, there is the need of high amounts of example data. 
-This data is of numeric type and different tabular data sources need to be semantically consolidated.
+Operating wind farms involves high costs for maintaining the wind
+turbines. Predicting potential failures and optimizing the maintenance
+operations to minimize the costs can be business critical. For
+establishing predictive maintenance, we need large amounts of high
+qualitative training data, which can be achieved by consolidating
+different wind farm measurement sources. This use case provides
+different kinds of tabular data, where consolidation needs semantic
+harmonization of data.
 
-### The Legal Dataspace
-Analyzing documents in the legal domain is a highly challenging task because of the complexity of legal information. 
-We provide recommender services for insights into corpora of legal documents 
-that identify the meaning via a legal knowledge graph, NLP analysis and semantic annotations.
+\subsection{Energy and Legal
+Dataspace}\label{energy-and-legal-dataspace}
 
-## The Problem
-"Standard" dataspace solutions provide access to data and metadata descriptions, but they do not typically address the data integration problem. 
-If a consumer wants to consume datasets from different providers in an integrated way, 
-they need to harmonize the data to some common model, convert it, store it in a database, 
-and potentially implement entity linking (correlation of different records that are about the same real-world thing) and data fusion.
+Beyond the UNDERPIN project as the main use case provider, we also
+consider use cases from other projects. We present 2 examples from the
+DataBri-X project, which work on different content types of data and
+have different usage scenarios. We considered these as well when
+designing our solution to make it applicable to a broader range of use
+cases.
 
-Above we gave just 4 examples to illustrate the heterogeneity of data in the context of different use cases. 
-We aim to develop solutions that not only fulfill the dataspace needs, 
-but also improve the situation regarding semantic interoperability by providing services that can be easily leveraged to implement such use cases.
+\subsubsection{The Energy Dataspace}\label{the-energy-dataspace}
 
-# Approach and IT-Solution
-To tackle this problem, we introduce Semantic Web standards and technologies to model and process data. 
-Vocabularies and ontologies (expressed using RDFS, OWL and SKOS) can represent data with a clear semantics based on standards. 
-We leverage GraphDB and PoolParty as two software components that implement these standards to process the data.
+In this use case, we run simulations for energy communities, which
+predict the behavior of the energy grid. To provide precise predictions,
+there is the need of high amounts of example data. This data is of
+numeric type and different tabular data sources need to be semantically
+consolidated.
 
-## IDS Vocabulary Hub
-In this paper, we focus on the IDSA approach to dataspaces. 
-IDSA defines the architecture of dataspaces in the reference architecture model IDS-RAM (current version 4). 
-One of the defined architectural components, the Vocabulary Hub, has the role of providing vocabularies to annotate and describe data assets and services.
-These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
-The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions. 
-It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies 
-and thereby provides semantic interoperability in a standardized and machine-readable way.
+\subsubsection{The Legal Dataspace}\label{the-legal-dataspace}
 
-## GraphDB Graph Database
+Analyzing documents in the legal domain is a highly challenging task
+because of the complexity of legal information. We provide recommender
+services for insights into corpora of legal documents that identify the
+meaning via a legal knowledge graph, NLP analysis and semantic
+annotations.
+
+\subsection{The Problem}\label{the-problem}
+
+``Standard'' dataspace solutions provide access to data and metadata
+descriptions, but they do not typically address the data integration
+problem. If a consumer wants to consume datasets from different
+providers in an integrated way, they need to harmonize the data to some
+common model, convert it, store it in a database, and potentially
+implement entity linking (correlation of different records that are
+about the same real-world thing) and data fusion.
+
+Above we gave just 4 examples to illustrate the heterogeneity of data in
+the context of different use cases. We aim to develop solutions that not
+only fulfill the dataspace needs, but also improve the situation
+regarding semantic interoperability by providing services that can be
+easily leveraged to implement such use cases.
+
+\section{Approach and IT-Solution}\label{approach-and-it-solution}
+
+To tackle this problem, we introduce Semantic Web standards and
+technologies to model and process data. Vocabularies and ontologies
+(expressed using RDFS, OWL and SKOS) can represent data with a clear
+semantics based on standards. We leverage GraphDB and PoolParty as two
+software components that implement these standards to process the data.
+
+\subsection{IDS Vocabulary Hub}\label{ids-vocabulary-hub}
+
+In this paper, we focus on the IDSA approach to dataspaces. IDSA defines
+the architecture of dataspaces in the reference architecture model
+IDS-RAM (current version 4). One of the defined architectural
+components, the Vocabulary Hub, has the role of providing vocabularies
+to annotate and describe data assets and services. These vocabularies
+form a common language to enable semantic interoperability among the
+dataspace participants. The Vocabulary Hub is responsible for resolving
+annotations of data assets and for providing details in the form of
+standardized semantic descriptions. It supports the use of Semantic Web
+standards for defining and representing RDF vocabularies and ontologies
+and thereby provides semantic interoperability in a standardized and
+machine-readable way.
+
+\subsection{GraphDB Graph Database}\label{graphdb-graph-database}
+
 TODO
 
-## PoolParty Semantic Suite
-PoolParty Semantic Suite is a semantic middleware platform based on W3C standards, specifically the Semantic Web, which provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.
+\subsection{PoolParty Semantic Suite}\label{poolparty-semantic-suite}
 
-At the core of PoolParty is the Thesaurus Manager, which can be used to create and maintain RDF-based vocabularies, including ontologies, schemas and different kinds of knowledge models.
-These vocabularies can be applied for various use cases, from bridging data silos in enterprises by implementing data consolidation via the vocabularies, to annotating documents with semantic annotations to implement semantic search and recommendations which provide insights into data.
+PoolParty Semantic Suite is a semantic middleware platform based on W3C
+standards, specifically the Semantic Web, which provides a wide variety
+of functionality in the area of knowledge graph management, graph-based
+NLP, semantic search and recommender systems.
 
-PoolParty is well suited to form the basis of a vocabulary management system within a dataspace, 
-effectively implementing not only the Vocabulary Hub role, 
-but also providing additional services that improve the expressiveness of common vocabularies and automate semantic interoperability.
+At the core of PoolParty is the Thesaurus Manager, which can be used to
+create and maintain RDF-based vocabularies, including ontologies,
+schemas and different kinds of knowledge models. These vocabularies can
+be applied for various use cases, from bridging data silos in
+enterprises by implementing data consolidation via the vocabularies, to
+annotating documents with semantic annotations to implement semantic
+search and recommendations which provide insights into data.
 
-## The GraphDB Poolparty Integration
-The integration of GraphDB and PoolParty leverages synergies of a high-performance graph database with a platform for knowledge graph management.
+PoolParty is well suited to form the basis of a vocabulary management
+system within a dataspace, effectively implementing not only the
+Vocabulary Hub role, but also providing additional services that improve
+the expressiveness of common vocabularies and automate semantic
+interoperability.
 
-Both GraphDB and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL.
-They therefore seamlessly integrate with each other, but also with other services and components in dataspaces which support these machine-readable standards for semantic interoperability.
+\subsection{The GraphDB Poolparty
+Integration}\label{the-graphdb-poolparty-integration}
 
-# A Semantic Solution and its Benefits
-When harmonizing data, we need to provide a method to automatically map and process it, 
-so this can be done efficiently on large datasets and/or regular updates in dataspaces. 
-The benefit provided should support the ML training in such a way that the consolidated dataset is exposed as a uniform training dataset.
+The integration of GraphDB and PoolParty leverages synergies of a
+high-performance graph database with a platform for knowledge graph
+management.
 
-## Semantic Layer Approach
-The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
-We aim to go one step further and envision a Semantic Layer for dataspaces, 
-which provides vocabulary-based services for advanced and automated semantic descriptions of metadata and data. 
-The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.
+Both GraphDB and PoolParty are based on W3C recommendations like RDF,
+OWL and SPARQL. They therefore seamlessly integrate with each other, but
+also with other services and components in dataspaces which support
+these machine-readable standards for semantic interoperability.
 
-In the following, we discuss 
-i) the services provided for the Vocabulary Hub component and 
-ii) additional services as Vocabulary Hub extensions that implement a Semantic Layer for dataspaces.
+\section{A Semantic Solution and its
+Benefits}\label{a-semantic-solution-and-its-benefits}
 
-Vocabulary Hub Services:
-- RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
-- Scalable vocabulary storage: cluster implementation for higher throughput and a fault tolerant architecture.
-- Knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs. 
-  Vocabularies can be created and edited based on standards like RDFS, OWL and SKOS, and existing vocabularies can be reused and combined to fit specific dataspace needs.
-- SPARQL service for semantic lookups: dereference URIs describing data assets to retrieve a semantic description of the entity from the graph database.
+When harmonizing data, we need to provide a method to automatically map
+and process it, so this can be done efficiently on large datasets and/or
+regular updates in dataspaces. The benefit provided should support the
+ML training in such a way that the consolidated dataset is exposed as a
+uniform training dataset.
 
-Semantic Layer Services support both unstructured and structured data assets:
-- Unstructured data: semantic annotation services for documents based on concept tagging. 
-  PoolParty’s KG-based NLP primarily uses SKOS taxonomies to identify concepts in documents, 
-  but can be extended with OWL ontology elements (classes, properties) for increased expressiveness.
-- Structured data: we support structured data harmonization methods, including semantic mapping and linking SKOS concepts
-  - Ontology-Based Data Access (OBDA) is used for virtualization of relational data, i.e. accessing relational data as semantic data. 
-    It is well suited for semantic querying of large amounts of data, while keeping the original relational data in place (NoETL approach). 
-    Ontologies are leveraged to construct the OBDA mapping; 
-    they are governed via the Semantic Layer and can harmonize data assets of various formats using a precise mapping. 
-    Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
-  - RDB to RDF Mapping Language (R2RML) [@souripriyadasR2RMLRDBRDF2012]
-    and RDF Mapping Language (RML) [@RMLIntroduction2024]
-    are intended for data transformation processes within a dataspace, 
-    where the original source is used as a basis to create an RDF representation (ETL approach).
-  - GraphDB supports both OBDA and R2RML through its integration of ONTOP [@bagosiOntopFrameworkOntology2014]. 
-    The same mappings can be used for NoETL virtual data access, or for ETL (to materialize the semantic data).
-  - CSV on the Web (CSVW) [@swirrlCSVWStandards2024] is a set of W3C standards 
-    that allow detailed semantic description of tabular data using a JSON "manifest" file.
-  - SKOS concepts are used for structured data in several ways. 
-    Concepts can be directly added to RDF data via semantic links (properties from an ontology), 
-    they can be used to annotate specific literals of an RDF graph, 
-    or they can be used to associate documents with (specific parts of) an RDF graph.
-- Inference services: we provide metadata expansion services based on interlinked vocabularies in combination with ontologies. 
-  This inference approach can be used with semantic tagging for unstructured data, as well as the structured data approaches.
-  - Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inference. 
-    The query can be configured to work for various scenarios.
-- Vocabulary crosswalks are a data modeling approach using interlinks to relate:
-  - Different vocabularies from similar or the same domain that cover different areas or aspects of the domain. 
-    By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand queries and annotations by traversing the interlinked graph.
-  - Different versions of the same vocabulary to automatically translate between those versions.
-- Metadata inference based on interlinked vocabularies is similar to crosswalks, 
-  but intended to represent different levels of abstraction, 
-  so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa. 
-  For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go into more detail.
+\subsection{Semantic Layer Approach}\label{semantic-layer-approach}
 
-## Software Architecture
+The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic
+building block to achieve semantic interoperability in dataspaces by
+provisioning common vocabularies. We aim to go one step further and
+envision a Semantic Layer for dataspaces, which provides
+vocabulary-based services for advanced and automated semantic
+descriptions of metadata and data. The Semantic Layer is implemented as
+services based on the GraphDB and PoolParty and provided as a one
+seamlessly integrated component.
+
+In the following, we discuss i) the services provided for the Vocabulary
+Hub component and ii) additional services as Vocabulary Hub extensions
+that implement a Semantic Layer for dataspaces.
+
+Vocabulary Hub Services: - RDF based: both GraphDB and PoolParty store
+and manage data based on the RDF data model. RDF is the basic building
+stone and underlying vocabulary definitions. - Scalable vocabulary
+storage: cluster implementation for higher throughput and a fault
+tolerant architecture. - Knowledge graph management: easy management of
+vocabularies using a rich web user interface and integration via Web
+APIs. Vocabularies can be created and edited based on standards like
+RDFS, OWL and SKOS, and existing vocabularies can be reused and combined
+to fit specific dataspace needs. - SPARQL service for semantic lookups:
+dereference URIs describing data assets to retrieve a semantic
+description of the entity from the graph database.
+
+Semantic Layer Services support both unstructured and structured data
+assets: - Unstructured data: semantic annotation services for documents
+based on concept tagging. PoolParty's KG-based NLP primarily uses SKOS
+taxonomies to identify concepts in documents, but can be extended with
+OWL ontology elements (classes, properties) for increased
+expressiveness. - Structured data: we support structured data
+harmonization methods, including semantic mapping and linking SKOS
+concepts - Ontology-Based Data Access (OBDA) is used for virtualization
+of relational data, i.e.~accessing relational data as semantic data. It
+is well suited for semantic querying of large amounts of data, while
+keeping the original relational data in place (NoETL approach).
+Ontologies are leveraged to construct the OBDA mapping; they are
+governed via the Semantic Layer and can harmonize data assets of various
+formats using a precise mapping. Furthermore, it can be used to
+establish a fine-grained access control, thereby deciding for individual
+data elements if they should be exposed ot not. - RDB to RDF Mapping
+Language (R2RML) \citep{souripriyadasR2RMLRDBRDF2012} and RDF Mapping
+Language (RML) \citep{RMLIntroduction2024} are intended for data
+transformation processes within a dataspace, where the original source
+is used as a basis to create an RDF representation (ETL approach). -
+GraphDB supports both OBDA and R2RML through its integration of ONTOP
+\citep{bagosiOntopFrameworkOntology2014}. The same mappings can be used
+for NoETL virtual data access, or for ETL (to materialize the semantic
+data). - CSV on the Web (CSVW) \citep{swirrlCSVWStandards2024} is a set
+of W3C standards that allow detailed semantic description of tabular
+data using a JSON ``manifest'' file. - SKOS concepts are used for
+structured data in several ways. Concepts can be directly added to RDF
+data via semantic links (properties from an ontology), they can be used
+to annotate specific literals of an RDF graph, or they can be used to
+associate documents with (specific parts of) an RDF graph. - Inference
+services: we provide metadata expansion services based on interlinked
+vocabularies in combination with ontologies. This inference approach can
+be used with semantic tagging for unstructured data, as well as the
+structured data approaches. - Inference tagging is provided by a
+PoolParty tagging service in combination with an expansion query for
+implementing the inference. The query can be configured to work for
+various scenarios. - Vocabulary crosswalks are a data modeling approach
+using interlinks to relate: - Different vocabularies from similar or the
+same domain that cover different areas or aspects of the domain. By
+relating concepts with each other using 1:1, 1:n and m:n relations, we
+can automatically expand queries and annotations by traversing the
+interlinked graph. - Different versions of the same vocabulary to
+automatically translate between those versions. - Metadata inference
+based on interlinked vocabularies is similar to crosswalks, but intended
+to represent different levels of abstraction, so that we are able to
+infer from a more general vocabulary to a more specific vocabulary and
+vice versa. For example, we can start with DCAT themes vocabularies and
+link to further vocabularies, which are more specific for certain areas
+or which go into more detail.
+
+\subsection{Software Architecture}\label{software-architecture}
 
@@
 
-![](img/architecture.png)
+\includegraphics{img/architecture.png}
 
-- A Data Provider sends data through the Connector, 
-  and at the same time provides basic metadata (DCAT)
-- Selected data can be stored in Database(s) managed by the dataspace, 
+\begin{itemize}
+\tightlist
+\item
+  A Data Provider sends data through the Connector, and at the same time
+  provides basic metadata (DCAT)
+\item
+  Selected data can be stored in Database(s) managed by the dataspace,
   for easier access and querying.
-  - For large-volume time series data, we recommend to use InfluxDB, PostgreSQL, Google BigQuery, etc.
-  - Of course, access control is very important here. 
-    We'll leverage GraphDB's [Fine-grained Access Control](https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html), 
-    but access control in the time series database is also important.
-  - Depending on Provider preferences and agreements, 
-    other data may stay at the source, 
-    and be delivered to the Consumer only dynamically through the Connector.
-- The Vocabulary Hub stores relevant semantic assets: ontologies and thesauri. 
-  These are added based on relevant use cases and datasets, following a dataspace governance process.
-- The Vocabulary Hub also stores mapping assets, such as OBDA or R2RML mappings and CSVW manifests.
-- PoolParty Semantic Integrator is used to edit, view and manage the semantic assets, 
-  and to bind mapping assets to incoming data.
-- For virtual access to relational data, ONTOP is used to translate SPARQL to SQL dynamically 
-  and to convert returned data to SPARQL result format.
-- The Data Consumer benefits from:
-  - More powerful dataset discovery by using richer semantic queries, e.g.
-    - "Give me all time series related to compressor C123" (equipment instance)
-    - "Give me all time series related to temperature" (measured quantity)
-    - "Give me all time series of bearings" (equipment kind or machine part)
-  - Harmonized querying of datasets from multiple providers.
 
-# Conclusion and Future Work
-We presented an approach for dynamically binding datasets to semantic descriptions in a dataspace's Vocabulary Hub, 
-therefore facilitating data harmonization and easier data consumption. 
-We have implemented a first prototype of our approach, which builds on the integration of GraphDB and PoolParty. 
-As next steps, we will bring this solution into use cases and into broader discussion to gain insight and develop it further.
+  \begin{itemize}
+  \tightlist
+  \item
+    For large-volume time series data, we recommend to use InfluxDB,
+    PostgreSQL, Google BigQuery, etc.
+  \item
+    Of course, access control is very important here. We'll leverage
+    GraphDB's
+    \href{https://graphdb.ontotext.com/documentation/10.6/fine-grained-access-control.html}{Fine-grained
+    Access Control}, but access control in the time series database is
+    also important.
+  \item
+    Depending on Provider preferences and agreements, other data may
+    stay at the source, and be delivered to the Consumer only
+    dynamically through the Connector.
+  \end{itemize}
+\item
+  The Vocabulary Hub stores relevant semantic assets: ontologies and
+  thesauri. These are added based on relevant use cases and datasets,
+  following a dataspace governance process.
+\item
+  The Vocabulary Hub also stores mapping assets, such as OBDA or R2RML
+  mappings and CSVW manifests.
+\item
+  PoolParty Semantic Integrator is used to edit, view and manage the
+  semantic assets, and to bind mapping assets to incoming data.
+\item
+  For virtual access to relational data, ONTOP is used to translate
+  SPARQL to SQL dynamically and to convert returned data to SPARQL
+  result format.
+\item
+  The Data Consumer benefits from:
 
-The next steps include:
-- Bringing it into actual dataspaces in practice, representing and integrating datasets from different dataspace participants.
-- Implementing use cases for richer discoverability, harmonized querying and support for different content types of structured and unstructured data.
-- Explore how ML can benefit regarding quality in practice when providing consolidated and cleaned data via dataspaces.
-- Discuss how we can extend the IDS-RAM with services to improve the support for semantic interoperability provided by our solution.
+  \begin{itemize}
+  \tightlist
+  \item
+    More powerful dataset discovery by using richer semantic queries,
+    e.g.
 
-# Acknowledgements
-This work is partially supported by the Digital Europe programme project [UNDERPIN](https://underpinproject.eu/) (grant agreement 101123179)
+    \begin{itemize}
+    \tightlist
+    \item
+      ``Give me all time series related to compressor C123'' (equipment
+      instance)
+    \item
+      ``Give me all time series related to temperature'' (measured
+      quantity)
+    \item
+      ``Give me all time series of bearings'' (equipment kind or machine
+      part)
+    \end{itemize}
+  \item
+    Harmonized querying of datasets from multiple providers.
+  \end{itemize}
+\end{itemize}
 
-# References
+\section{Conclusion and Future Work}\label{conclusion-and-future-work}
+
+We presented an approach for dynamically binding datasets to semantic
+descriptions in a dataspace's Vocabulary Hub, therefore facilitating
+data harmonization and easier data consumption. We have implemented a
+first prototype of our approach, which builds on the integration of
+GraphDB and PoolParty. As next steps, we will bring this solution into
+use cases and into broader discussion to gain insight and develop it
+further.
+
+The next steps include: - Bringing it into actual dataspaces in
+practice, representing and integrating datasets from different dataspace
+participants. - Implementing use cases for richer discoverability,
+harmonized querying and support for different content types of
+structured and unstructured data. - Explore how ML can benefit regarding
+quality in practice when providing consolidated and cleaned data via
+dataspaces. - Discuss how we can extend the IDS-RAM with services to
+improve the support for semantic interoperability provided by our
+solution.
+
+\section{Acknowledgements}\label{acknowledgements}
+
+This work is partially supported by the Digital Europe programme project
+\href{https://underpinproject.eu/}{UNDERPIN} (grant agreement 101123179)
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bibliography.bib}
+
+%% begin pandoc before-bib
+%% end pandoc before-bib
+%% begin pandoc biblio
+%% end pandoc biblio
+%% begin pandoc include-after
+%% end pandoc include-after
+%% begin pandoc after-body
+%% end pandoc after-body
+
+\end{document}
+\endinput
+%%
+%% End of file `sample-manuscript.tex'.

### 2024-06-15 EBDVF-2024-elsevier/paper.qmd
updated EBDVF paper using elsevier template

--- 2024-06-14 EBDVF-2024/paper.qmd
@@
 title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
-
+author:
+  - name: Robert David
+    email: robert.david@semantic-web.com
+    orcid: 0000-0002-3244-5341
+    affiliations: 
+        - id: swc
+          name: The Semantic Web Company
+          country: Austria
+  - name: Petar Ivanov
+    email: petar.ivanov@ontotext.com
+    orcid: 0000-0001-8448-1005
+    affiliations: 
+        - id: onto
+          name: Ontotext
+          country: Bulgaria
+  - name: Vladimir Alexiev
+    email: vladimir.alexiev@ontotext.com
+    orcid: 0000-0001-7508-7428
+    affiliations:
+        - ref: onto
 abstract: |
@@
   with applications to refinery and wind farm data.
-
-author:
-  - name: Robert David
-    email: robert.david@semantic-web.com
-    orcid: 0000-0002-3244-5341
-    affiliation: 
-     name: The Semantic Web Company
-     country: Austria
-  - name: Vladimir Alexiev
-    email: vladimir.alexiev@ontotext.com
-    orcid: 0000-0001-7508-7428
-  - name: Petar Ivanov
-    email: petar.ivanov@ontotext.com
-    orcid: 0000-0001-8448-1005
-    affiliation: 
-      name: Ontotext
-      country: Bulgaria
-acm-metadata:
 keywords: 
@@
   - windfarms
-  final: true
-  # acmart-options: sigconf, review
-  # copyright-year: 2024
-  # acm-year: 2024
-  copyright: acmcopyright
-  # doi: 
-  # conference-acronym: 
-  # conference-name: 
-  # conference-date: 
-  # conference-location: 
-  # isbn:
-
+date: last-modified
+bibliography: bibliography.bib
 format:
-  acm-pdf:
+  elsevier-pdf:
     keep-tex: true
-bibliography: bibliography.bib
-
+    journal:
+      name: Workshop on Semantic Interoperability in Data Spaces 
+      formatting: preprint
+      model: 3p
+      cite-style: super
 ---
@@
 Examples are the well-known mobility dataspace and use cases like energy communities for running simulations. 
-To cope with the challenge of data sharing, Dataspace approaches like the IDS RAM were introduced to not only solve the technical aspects of data sharing, 
+To cope with the challenge of data sharing, Dataspace approaches like the IDS Reference Architecture Model (IDS RAM) were introduced to not only solve the technical aspects of data sharing, 
 but also to establish methods for data sovereignty so that dataspace participants can determine themselves how and when others can make use of their data.
@@
 For making use of shared data, interoperability is a crucial factor. 
-While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability. 
+While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability [@woutervandenbergSemanticInteroperabilityData2024]. 
 IDS RAM is firmly based on semantic metadata for all aspects and actors of a dataspace. 
-Furthermore, it can leverage semantic interoperability approaches through the Vocabulary Hub 
+Furthermore, it can leverage semantic interoperability approaches through the Vocabulary Hub [@IDSARAMVocabulary2024]
 that stores ontologies and semantic thesauri for data descriptions and can dereference URIs to provide semantic details of data assets.
@@
 which provides even more services to easily and semi-automatically manage semantic interoperability. 
-Thus we raise the role of the Vocabulary Hub to not only store semantic assets, 
+Thus we raise the role of the Vocabulary Hub [@stornebrinkmichielVocabularyHubSemantic2023] to not only store semantic assets, 
 but also to bind semantic descriptions dynamically to incoming data, 
@@
 ## GraphDB Graph Database
-TODO
+Ontotext GraphDB is a highly efficient, scalable and robust graph database with RDF and SPARQL support.
+It also offers data virtualization capabilities which enable direct access using SPARQL to relational databases, eliminating the need to replicate data.
+It can perform semantic inferencing at scale, which enables deriving new semantic facts from existing facts.
+
+GraphDB handles massive loads, queries, and inferencing in real time.
+Reasoning and query evaluation are performed over a persistent storage layer. Loading, reasoning, and query evaluation are executed extremely fast even against huge ontologies and knowledge bases.
+According to the [LDBC Semantic Publishing Benchmark](https://ldbcouncil.org/benchmarks/spb/), it is one of the most scalable OWL repositories currently available.
 
@@
 Vocabulary Hub Services:
+
 - RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
@@
 Semantic Layer Services support both unstructured and structured data assets:
+
 - Unstructured data: semantic annotation services for documents based on concept tagging. 
@@
     Ontologies are leveraged to construct the OBDA mapping; 
-    they are governed via the Semantic Layer and can harmonize data assets of various formats using a precise mapping. 
+    they are governed via the Semantic Layer and can harmonize data assets of various formats using a precise mapping [@andreselAdaptingOntologybasedData2024]. 
     Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
@@
 The next steps include:
+
 - Bringing it into actual dataspaces in practice, representing and integrating datasets from different dataspace participants.
@@
 # Acknowledgements
-This work is partially supported by the Digital Europe programme project [UNDERPIN](https://underpinproject.eu/) (grant agreement 101123179)
+This work is partially supported by the Digital Europe programme project [UNDERPIN](https://underpinproject.eu/) (grant agreement 101123179).
 
-# References
+# References {-}

### 2024-06-15 EBDVF-2024-elsevier/paper.tex
updated EBDVF paper using elsevier template

--- 2024-06-14 EBDVF-2024/paper.tex
@@
-\documentclass[manuscript,screen]{acmart}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  super,
+  preprint,
+  3p]{elsarticle}
 
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 \makeatother
-
-%%
-%% This is file `sample-manuscript.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `manuscript')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-manuscript.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass command.
-
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-
-%% PANDOC PREAMBLE BEGINS
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{5}
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
@@
 
-\definecolor{mypink}{RGB}{219, 48, 122}
 \makeatletter
@@
 \makeatother
-%% PANDOC PREAMBLE ENDS
+\journal{Workshop on Semantic Interoperability in Data Spaces}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-num}
+\usepackage{bookmark}
 
-\setlength{\parindent}{10pt}
-\setlength{\parskip}{0pt}
-
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
 \hypersetup{
   pdftitle={Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces},
-  pdfauthor={Robert David; Vladimir Alexiev; Petar Ivanov},
+  pdfauthor={Robert David; Petar Ivanov; Vladimir Alexiev},
+  pdfkeywords={dataspaces, semantic interoperability, semantic
+technologies, ontologies, vocabulary hub, oil and gas, renewable
+energy, refineries, windfarms},
   colorlinks=true,
@@
   citecolor={Blue},
-  urlcolor={red},
-  pdfcreator={LaTeX via pandoc, via quarto}}
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
 
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{}
-\acmYear{}
-\acmDOI{}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[]{}{}{}
-\acmPrice{}
-\acmISBN{}
-
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-%%\citestyle{acmauthoryear}
-
-%% end of the preamble, start of the body of the document source.
+\setlength{\parindent}{6pt}
 \begin{document}
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
+\begin{frontmatter}
 \title{Raising the Role of Vocabulary Hubs for Semantic Data
 Interoperability in Dataspaces}
+\author[1]{Robert David%
+%
+}
+ \ead{robert.david@semantic-web.com} 
+\author[2]{Petar Ivanov%
+%
+}
+ \ead{petar.ivanov@ontotext.com} 
+\author[2]{Vladimir Alexiev%
+%
+}
+ \ead{vladimir.alexiev@ontotext.com}
 
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
+\affiliation[1]{organization={The Semantic Web
+Company},country={Austria},countrysep={,},postcodesep={}}
+\affiliation[2]{organization={Ontotext},country={Bulgaria},countrysep={,},postcodesep={}}
 
-  \author{Robert David}
-  \orcid{0000-0002-3244-5341}
-            \affiliation{%
-                  \institution{The Semantic Web Company}
-                                                  \country{Austria}
-                      }
-        \author{Vladimir Alexiev}
-  \orcid{0000-0001-7508-7428}
-    \author{Petar Ivanov}
-  \orcid{0000-0001-8448-1005}
-            \affiliation{%
-                  \institution{Ontotext}
-                                                  \country{Bulgaria}
-                      }
+\cortext[cor1]{Corresponding author}
 
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-%\renewcommand{\shortauthors}{Trovato et al.}
-%%  
-%% The abstract is a short summary of the work to be presented in the
-%% article.
 \begin{abstract}
@@
 
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{dataspaces, semantic interoperability, semantic
-technologies, ontologies, vocabulary hub, oil and gas, renewable
-energy, refineries, windfarms}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
-
-\setlength{\parskip}{-0.1pt}
+\begin{keyword}
+    dataspaces \sep semantic interoperability \sep semantic
+technologies \sep ontologies \sep vocabulary hub \sep oil and
+gas \sep renewable energy \sep refineries \sep 
+    windfarms
+\end{keyword}
+\end{frontmatter}
 
@@
 communities for running simulations. To cope with the challenge of data
-sharing, Dataspace approaches like the IDS RAM were introduced to not
-only solve the technical aspects of data sharing, but also to establish
-methods for data sovereignty so that dataspace participants can
-determine themselves how and when others can make use of their data.
+sharing, Dataspace approaches like the IDS Reference Architecture Model
+(IDS RAM) were introduced to not only solve the technical aspects of
+data sharing, but also to establish methods for data sovereignty so that
+dataspace participants can determine themselves how and when others can
+make use of their data.
 
@@
 While syntactic data exchange is well covered by clearly defined data
-formats, there is still the open challenge of semantic interoperability.
-IDS RAM is firmly based on semantic metadata for all aspects and actors
-of a dataspace. Furthermore, it can leverage semantic interoperability
-approaches through the Vocabulary Hub that stores ontologies and
-semantic thesauri for data descriptions and can dereference URIs to
-provide semantic details of data assets.
+formats, there is still the open challenge of semantic interoperability
+\citep{woutervandenbergSemanticInteroperabilityData2024}. IDS RAM is
+firmly based on semantic metadata for all aspects and actors of a
+dataspace. Furthermore, it can leverage semantic interoperability
+approaches through the Vocabulary Hub \citep{IDSARAMVocabulary2024} that
+stores ontologies and semantic thesauri for data descriptions and can
+dereference URIs to provide semantic details of data assets.
 
@@
 semantic interoperability. Thus we raise the role of the Vocabulary Hub
-to not only store semantic assets, but also to bind semantic
-descriptions dynamically to incoming data, therefore enabling richer
-discovery and semantic data integration between datasets.
+\citep{stornebrinkmichielVocabularyHubSemantic2023} to not only store
+semantic assets, but also to bind semantic descriptions dynamically to
+incoming data, therefore enabling richer discovery and semantic data
+integration between datasets.
 
@@
 
-TODO
+Ontotext GraphDB is a highly efficient, scalable and robust graph
+database with RDF and SPARQL support. It also offers data virtualization
+capabilities which enable direct access using SPARQL to relational
+databases, eliminating the need to replicate data. It can perform
+semantic inferencing at scale, which enables deriving new semantic facts
+from existing facts.
+
+GraphDB handles massive loads, queries, and inferencing in real time.
+Reasoning and query evaluation are performed over a persistent storage
+layer. Loading, reasoning, and query evaluation are executed extremely
+fast even against huge ontologies and knowledge bases. According to the
+\href{https://ldbcouncil.org/benchmarks/spb/}{LDBC Semantic Publishing
+Benchmark}, it is one of the most scalable OWL repositories currently
+available.
 
@@
 
-Vocabulary Hub Services: - RDF based: both GraphDB and PoolParty store
-and manage data based on the RDF data model. RDF is the basic building
-stone and underlying vocabulary definitions. - Scalable vocabulary
-storage: cluster implementation for higher throughput and a fault
-tolerant architecture. - Knowledge graph management: easy management of
-vocabularies using a rich web user interface and integration via Web
-APIs. Vocabularies can be created and edited based on standards like
-RDFS, OWL and SKOS, and existing vocabularies can be reused and combined
-to fit specific dataspace needs. - SPARQL service for semantic lookups:
-dereference URIs describing data assets to retrieve a semantic
-description of the entity from the graph database.
+Vocabulary Hub Services:
+
+\begin{itemize}
+\tightlist
+\item
+  RDF based: both GraphDB and PoolParty store and manage data based on
+  the RDF data model. RDF is the basic building stone and underlying
+  vocabulary definitions.
+\item
+  Scalable vocabulary storage: cluster implementation for higher
+  throughput and a fault tolerant architecture.
+\item
+  Knowledge graph management: easy management of vocabularies using a
+  rich web user interface and integration via Web APIs. Vocabularies can
+  be created and edited based on standards like RDFS, OWL and SKOS, and
+  existing vocabularies can be reused and combined to fit specific
+  dataspace needs.
+\item
+  SPARQL service for semantic lookups: dereference URIs describing data
+  assets to retrieve a semantic description of the entity from the graph
+  database.
+\end{itemize}
 
 Semantic Layer Services support both unstructured and structured data
-assets: - Unstructured data: semantic annotation services for documents
-based on concept tagging. PoolParty's KG-based NLP primarily uses SKOS
+assets:
+
+\begin{itemize}
+\tightlist
+\item
+  Unstructured data: semantic annotation services for documents based on
+  concept tagging. PoolParty's KG-based NLP primarily uses SKOS
   taxonomies to identify concepts in documents, but can be extended with
   OWL ontology elements (classes, properties) for increased
-expressiveness. - Structured data: we support structured data
-harmonization methods, including semantic mapping and linking SKOS
-concepts - Ontology-Based Data Access (OBDA) is used for virtualization
-of relational data, i.e.~accessing relational data as semantic data. It
+  expressiveness.
+\item
+  Structured data: we support structured data harmonization methods,
+  including semantic mapping and linking SKOS concepts
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Ontology-Based Data Access (OBDA) is used for virtualization of
+    relational data, i.e.~accessing relational data as semantic data. It
     is well suited for semantic querying of large amounts of data, while
@@
     Ontologies are leveraged to construct the OBDA mapping; they are
-governed via the Semantic Layer and can harmonize data assets of various
-formats using a precise mapping. Furthermore, it can be used to
-establish a fine-grained access control, thereby deciding for individual
-data elements if they should be exposed ot not. - RDB to RDF Mapping
-Language (R2RML) \citep{souripriyadasR2RMLRDBRDF2012} and RDF Mapping
-Language (RML) \citep{RMLIntroduction2024} are intended for data
-transformation processes within a dataspace, where the original source
-is used as a basis to create an RDF representation (ETL approach). -
-GraphDB supports both OBDA and R2RML through its integration of ONTOP
-\citep{bagosiOntopFrameworkOntology2014}. The same mappings can be used
-for NoETL virtual data access, or for ETL (to materialize the semantic
-data). - CSV on the Web (CSVW) \citep{swirrlCSVWStandards2024} is a set
-of W3C standards that allow detailed semantic description of tabular
-data using a JSON ``manifest'' file. - SKOS concepts are used for
-structured data in several ways. Concepts can be directly added to RDF
-data via semantic links (properties from an ontology), they can be used
-to annotate specific literals of an RDF graph, or they can be used to
-associate documents with (specific parts of) an RDF graph. - Inference
-services: we provide metadata expansion services based on interlinked
-vocabularies in combination with ontologies. This inference approach can
-be used with semantic tagging for unstructured data, as well as the
-structured data approaches. - Inference tagging is provided by a
-PoolParty tagging service in combination with an expansion query for
-implementing the inference. The query can be configured to work for
-various scenarios. - Vocabulary crosswalks are a data modeling approach
-using interlinks to relate: - Different vocabularies from similar or the
-same domain that cover different areas or aspects of the domain. By
-relating concepts with each other using 1:1, 1:n and m:n relations, we
-can automatically expand queries and annotations by traversing the
-interlinked graph. - Different versions of the same vocabulary to
-automatically translate between those versions. - Metadata inference
-based on interlinked vocabularies is similar to crosswalks, but intended
-to represent different levels of abstraction, so that we are able to
-infer from a more general vocabulary to a more specific vocabulary and
-vice versa. For example, we can start with DCAT themes vocabularies and
-link to further vocabularies, which are more specific for certain areas
-or which go into more detail.
+    governed via the Semantic Layer and can harmonize data assets of
+    various formats using a precise mapping
+    \citep{andreselAdaptingOntologybasedData2024}. Furthermore, it can
+    be used to establish a fine-grained access control, thereby deciding
+    for individual data elements if they should be exposed ot not.
+  \item
+    RDB to RDF Mapping Language (R2RML)
+    \citep{souripriyadasR2RMLRDBRDF2012} and RDF Mapping Language (RML)
+    \citep{RMLIntroduction2024} are intended for data transformation
+    processes within a dataspace, where the original source is used as a
+    basis to create an RDF representation (ETL approach).
+  \item
+    GraphDB supports both OBDA and R2RML through its integration of
+    ONTOP \citep{bagosiOntopFrameworkOntology2014}. The same mappings
+    can be used for NoETL virtual data access, or for ETL (to
+    materialize the semantic data).
+  \item
+    CSV on the Web (CSVW) \citep{swirrlCSVWStandards2024} is a set of
+    W3C standards that allow detailed semantic description of tabular
+    data using a JSON ``manifest'' file.
+  \item
+    SKOS concepts are used for structured data in several ways. Concepts
+    can be directly added to RDF data via semantic links (properties
+    from an ontology), they can be used to annotate specific literals of
+    an RDF graph, or they can be used to associate documents with
+    (specific parts of) an RDF graph.
+  \end{itemize}
+\item
+  Inference services: we provide metadata expansion services based on
+  interlinked vocabularies in combination with ontologies. This
+  inference approach can be used with semantic tagging for unstructured
+  data, as well as the structured data approaches.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Inference tagging is provided by a PoolParty tagging service in
+    combination with an expansion query for implementing the inference.
+    The query can be configured to work for various scenarios.
+  \end{itemize}
+\item
+  Vocabulary crosswalks are a data modeling approach using interlinks to
+  relate:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Different vocabularies from similar or the same domain that cover
+    different areas or aspects of the domain. By relating concepts with
+    each other using 1:1, 1:n and m:n relations, we can automatically
+    expand queries and annotations by traversing the interlinked graph.
+  \item
+    Different versions of the same vocabulary to automatically translate
+    between those versions.
+  \end{itemize}
+\item
+  Metadata inference based on interlinked vocabularies is similar to
+  crosswalks, but intended to represent different levels of abstraction,
+  so that we are able to infer from a more general vocabulary to a more
+  specific vocabulary and vice versa. For example, we can start with
+  DCAT themes vocabularies and link to further vocabularies, which are
+  more specific for certain areas or which go into more detail.
+\end{itemize}
 
@@
 
-The next steps include: - Bringing it into actual dataspaces in
-practice, representing and integrating datasets from different dataspace
-participants. - Implementing use cases for richer discoverability,
-harmonized querying and support for different content types of
-structured and unstructured data. - Explore how ML can benefit regarding
-quality in practice when providing consolidated and cleaned data via
-dataspaces. - Discuss how we can extend the IDS-RAM with services to
-improve the support for semantic interoperability provided by our
-solution.
+The next steps include:
+
+\begin{itemize}
+\tightlist
+\item
+  Bringing it into actual dataspaces in practice, representing and
+  integrating datasets from different dataspace participants.
+\item
+  Implementing use cases for richer discoverability, harmonized querying
+  and support for different content types of structured and unstructured
+  data.
+\item
+  Explore how ML can benefit regarding quality in practice when
+  providing consolidated and cleaned data via dataspaces.
+\item
+  Discuss how we can extend the IDS-RAM with services to improve the
+  support for semantic interoperability provided by our solution.
+\end{itemize}
 
@@
 This work is partially supported by the Digital Europe programme project
-\href{https://underpinproject.eu/}{UNDERPIN} (grant agreement 101123179)
+\href{https://underpinproject.eu/}{UNDERPIN} (grant agreement
+101123179).
 
-\bibliographystyle{ACM-Reference-Format}
+\renewcommand\refname{References}
   \bibliography{bibliography.bib}
 
-%% begin pandoc before-bib
-%% end pandoc before-bib
-%% begin pandoc biblio
-%% end pandoc biblio
-%% begin pandoc include-after
-%% end pandoc include-after
-%% begin pandoc after-body
-%% end pandoc after-body
-
 \end{document}
-\endinput
-%%
-%% End of file `sample-manuscript.tex'.

### 2024-06-15 EBDVF-2024/paper.qmd
Add files via upload

@@
 
+The Vocabulary Hub is one of the core components of any dataspace not only those designed in accordance with IDS-RAM. 
+Other reference architecture models, such as GAIA-X architecture document 22.10 Release [@GaiaXArchitectureDocument2022] and Fraunhofer's Reference Architecture Model for the Industrial Data Space [@ReferenceArhictectureModel2017] also describe components with a similar role - the Federated Catalogue and Broker Service, respectively.
+Therefore, our approach designed around the IDS-RAM is adaptable and applicable to data spaces designed in accordance with other reference architecture models.
+
 ## GraphDB Graph Database
-TODO
+Ontotext GraphDB is a highly efficient, scalable and robust graph database with RDF and SPARQL support.
+It also offers data virtualization capabilities which enable direct access using SPARQL to relational databases, eliminating the need to replicate data.
+It can perform semantic inferencing at scale, which enables deriving new semantic facts from existing facts.
+
+GraphDB handles massive loads, queries, and inferencing in real time.
+Reasoning and query evaluation are performed over a persistent storage layer. Loading, reasoning, and query evaluation are executed extremely fast even against huge ontologies and knowledge bases.
+According to the [LDBC Semantic Publishing Benchmark](https://ldbcouncil.org/benchmarks/spb/), it is one of the most scalable OWL repositories currently available.
 
@@
 The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
-We aim to go one step further and envision a Semantic Layer for dataspaces, 
+Those vocabularies are to be expressed in RDF and by syntactically compatible with OWL to allow for expressive semantics. 
+The IDS-RAM does not require, but nonetheless allows for, extending the function of the Vocabulary hub towards providing ontology mappings [@FunctionalLayerIDSRAM2023b], which would enable dataspace connectors to automatically convert data between user-specific and standard data formats.
+Such automatic conversions are to be performed by Data Tranformation Apps by either the Data Consumers, the Data Providers, or both.
+
+We aim to achieve this through a Semantic Layer for dataspaces, 
 which provides vocabulary-based services for advanced and automated semantic descriptions of metadata and data. 
@@
 In the following, we discuss
+
 i) the services provided for the Vocabulary Hub component and 

### 2024-06-15 EBDVF-2024/paper.tex
Add files via upload

@@
 
+The Vocabulary Hub is one of the core components of any dataspace not
+only those designed in accordance with IDS-RAM. Other reference
+architecture models, such as GAIA-X architecture document 22.10 Release
+\citep{GaiaXArchitectureDocument2022} and Fraunhofer's Reference
+Architecture Model for the Industrial Data Space
+\citep{ReferenceArhictectureModel2017} also describe components with a
+similar role - the Federated Catalogue and Broker Service, respectively.
+Therefore, our approach designed around the IDS-RAM is adaptable and
+applicable to data spaces designed in accordance with other reference
+architecture models.
+
 \subsection{GraphDB Graph Database}\label{graphdb-graph-database}
 
-TODO
+Ontotext GraphDB is a highly efficient, scalable and robust graph
+database with RDF and SPARQL support. It also offers data virtualization
+capabilities which enable direct access using SPARQL to relational
+databases, eliminating the need to replicate data. It can perform
+semantic inferencing at scale, which enables deriving new semantic facts
+from existing facts.
+
+GraphDB handles massive loads, queries, and inferencing in real time.
+Reasoning and query evaluation are performed over a persistent storage
+layer. Loading, reasoning, and query evaluation are executed extremely
+fast even against huge ontologies and knowledge bases. According to the
+\href{https://ldbcouncil.org/benchmarks/spb/}{LDBC Semantic Publishing
+Benchmark}, it is one of the most scalable OWL repositories currently
+available.
 
@@
 building block to achieve semantic interoperability in dataspaces by
-provisioning common vocabularies. We aim to go one step further and
-envision a Semantic Layer for dataspaces, which provides
-vocabulary-based services for advanced and automated semantic
+provisioning common vocabularies. Those vocabularies are to be expressed
+in RDF and by syntactically compatible with OWL to allow for expressive
+semantics. The IDS-RAM does not require, but nonetheless allows for,
+extending the function of the Vocabulary hub towards providing ontology
+mappings \citep{FunctionalLayerIDSRAM2023b}, which would enable
+dataspace connectors to automatically convert data between user-specific
+and standard data formats. Such automatic conversions are to be
+performed by Data Tranformation Apps by either the Data Consumers, the
+Data Providers, or both.
+
+We aim to achieve this through a Semantic Layer for dataspaces, which
+provides vocabulary-based services for advanced and automated semantic
 descriptions of metadata and data. The Semantic Layer is implemented as
@@
 
-In the following, we discuss i) the services provided for the Vocabulary
-Hub component and ii) additional services as Vocabulary Hub extensions
-that implement a Semantic Layer for dataspaces.
+In the following, we discuss
 
-Vocabulary Hub Services: - RDF based: both GraphDB and PoolParty store
-and manage data based on the RDF data model. RDF is the basic building
-stone and underlying vocabulary definitions. - Scalable vocabulary
-storage: cluster implementation for higher throughput and a fault
-tolerant architecture. - Knowledge graph management: easy management of
-vocabularies using a rich web user interface and integration via Web
-APIs. Vocabularies can be created and edited based on standards like
-RDFS, OWL and SKOS, and existing vocabularies can be reused and combined
-to fit specific dataspace needs. - SPARQL service for semantic lookups:
-dereference URIs describing data assets to retrieve a semantic
-description of the entity from the graph database.
+\begin{enumerate}
+\def\labelenumi{\roman{enumi})}
+\tightlist
+\item
+  the services provided for the Vocabulary Hub component and
+\item
+  additional services as Vocabulary Hub extensions that implement a
+  Semantic Layer for dataspaces.
+\end{enumerate}
+
+Vocabulary Hub Services:
+
+\begin{itemize}
+\tightlist
+\item
+  RDF based: both GraphDB and PoolParty store and manage data based on
+  the RDF data model. RDF is the basic building stone and underlying
+  vocabulary definitions.
+\item
+  Scalable vocabulary storage: cluster implementation for higher
+  throughput and a fault tolerant architecture.
+\item
+  Knowledge graph management: easy management of vocabularies using a
+  rich web user interface and integration via Web APIs. Vocabularies can
+  be created and edited based on standards like RDFS, OWL and SKOS, and
+  existing vocabularies can be reused and combined to fit specific
+  dataspace needs.
+\item
+  SPARQL service for semantic lookups: dereference URIs describing data
+  assets to retrieve a semantic description of the entity from the graph
+  database.
+\end{itemize}
 
 Semantic Layer Services support both unstructured and structured data
-assets: - Unstructured data: semantic annotation services for documents
-based on concept tagging. PoolParty's KG-based NLP primarily uses SKOS
+assets:
+
+\begin{itemize}
+\tightlist
+\item
+  Unstructured data: semantic annotation services for documents based on
+  concept tagging. PoolParty's KG-based NLP primarily uses SKOS
   taxonomies to identify concepts in documents, but can be extended with
   OWL ontology elements (classes, properties) for increased
-expressiveness. - Structured data: we support structured data
-harmonization methods, including semantic mapping and linking SKOS
-concepts - Ontology-Based Data Access (OBDA) is used for virtualization
-of relational data, i.e.~accessing relational data as semantic data. It
+  expressiveness.
+\item
+  Structured data: we support structured data harmonization methods,
+  including semantic mapping and linking SKOS concepts
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Ontology-Based Data Access (OBDA) is used for virtualization of
+    relational data, i.e.~accessing relational data as semantic data. It
     is well suited for semantic querying of large amounts of data, while

### 2024-06-15 EBDVF-2024/paper.qmd
Update paper.qmd

@@
   For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
-  The Vocabulary Hub service (part of the IDSA RAM) provides a repository for ontologies and vocabularies.
+  The Vocabulary Hub service (part of the IDS RAM) provides a repository for ontologies and vocabularies.
   We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
@@
 In this paper, we focus on the IDSA approach to dataspaces. 
-IDSA defines the architecture of dataspaces in the reference architecture model IDS-RAM (current version 4). 
+IDSA defines the architecture of dataspaces in the reference architecture model IDS RAM (current version 4). 
 One of the defined architectural components, the Vocabulary Hub, has the role of providing vocabularies to annotate and describe data assets and services. 
@@
 
-The Vocabulary Hub is one of the core components of any dataspace not only those designed in accordance with IDS-RAM. 
+The Vocabulary Hub is one of the core components of any dataspace not only those designed in accordance with IDS RAM. 
 Other reference architecture models, such as GAIA-X architecture document 22.10 Release [@GaiaXArchitectureDocument2022] and Fraunhofer's Reference Architecture Model for the Industrial Data Space [@ReferenceArhictectureModel2017] also describe components with a similar role - the Federated Catalogue and Broker Service, respectively.
-Therefore, our approach designed around the IDS-RAM is adaptable and applicable to data spaces designed in accordance with other reference architecture models.
+Therefore, our approach designed around the IDS RAM is adaptable and applicable to data spaces designed in accordance with other reference architecture models.
 
@@
 ## Semantic Layer Approach
-The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
+The Vocabulary Hub, as IDS defines it in the IDS RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
 Those vocabularies are to be expressed in RDF and by syntactically compatible with OWL to allow for expressive semantics. 
-The IDS-RAM does not require, but nonetheless allows for, extending the function of the Vocabulary hub towards providing ontology mappings [@FunctionalLayerIDSRAM2023b], which would enable dataspace connectors to automatically convert data between user-specific and standard data formats.
+The IDS RAM does not require, but nonetheless allows for, extending the function of the Vocabulary hub towards providing ontology mappings [@FunctionalLayerIDSRAM2023b], which would enable dataspace connectors to automatically convert data between user-specific and standard data formats.
 Such automatic conversions are to be performed by Data Tranformation Apps by either the Data Consumers, the Data Providers, or both.
@@
 - Explore how ML can benefit regarding quality in practice when providing consolidated and cleaned data via dataspaces.
-- Discuss how we can extend the IDS-RAM with services to improve the support for semantic interoperability provided by our solution.
+- Discuss how we can extend the IDS RAM with services to improve the support for semantic interoperability provided by our solution.
 

### 2024-06-15 EBDVF-2024/paper.tex
Add files via upload

@@
 semantically integrated form. The Vocabulary Hub service (part of the
-IDSA RAM) provides a repository for ontologies and vocabularies. We
+IDS RAM) provides a repository for ontologies and vocabularies. We
 describe an approach of raising the role of the vocabulary hub to also
@@
 In this paper, we focus on the IDSA approach to dataspaces. IDSA defines
-the architecture of dataspaces in the reference architecture model
-IDS-RAM (current version 4). One of the defined architectural
-components, the Vocabulary Hub, has the role of providing vocabularies
-to annotate and describe data assets and services. These vocabularies
-form a common language to enable semantic interoperability among the
-dataspace participants. The Vocabulary Hub is responsible for resolving
+the architecture of dataspaces in the reference architecture model IDS
+RAM (current version 4). One of the defined architectural components,
+the Vocabulary Hub, has the role of providing vocabularies to annotate
+and describe data assets and services. These vocabularies form a common
+language to enable semantic interoperability among the dataspace
+participants. The Vocabulary Hub is responsible for resolving
 annotations of data assets and for providing details in the form of
@@
 The Vocabulary Hub is one of the core components of any dataspace not
-only those designed in accordance with IDS-RAM. Other reference
+only those designed in accordance with IDS RAM. Other reference
 architecture models, such as GAIA-X architecture document 22.10 Release
@@
 similar role - the Federated Catalogue and Broker Service, respectively.
-Therefore, our approach designed around the IDS-RAM is adaptable and
+Therefore, our approach designed around the IDS RAM is adaptable and
 applicable to data spaces designed in accordance with other reference
@@
 
-The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic
+The Vocabulary Hub, as IDS defines it in the IDS RAM, is a basic
 building block to achieve semantic interoperability in dataspaces by
@@
 in RDF and by syntactically compatible with OWL to allow for expressive
-semantics. The IDS-RAM does not require, but nonetheless allows for,
+semantics. The IDS RAM does not require, but nonetheless allows for,
 extending the function of the Vocabulary hub towards providing ontology
@@
 \item
-  Discuss how we can extend the IDS-RAM with services to improve the
+  Discuss how we can extend the IDS RAM with services to improve the
   support for semantic interoperability provided by our solution.

### 2024-06-16 EBDVF-2024/paper.qmd
edits

@@
     orcid: 0000-0002-3244-5341
+    corresponding: true
     affiliations:
@@
     orcid: 0000-0001-7508-7428
+    corresponding: true
     affiliations:
@@
   - windfarms
+journal:
+  highlights: # This makes a first page only with the Elsevier highlights: maybe we should remove it
+    - The Vocabulary Hub is used in a dataspace to hold semantic vocabularies (ontologies and thesauri).
+    - Proposes and implements a Semantic Layers architecture that allows binding semantic vocabularies to incoming datasets.
+    - Uses semantic mapping languages (OBDA, R2RML, CSVW) to harmonize datasets dynamically.
 date: last-modified
@@
 ## IDS Vocabulary Hub
-In this paper, we focus on the IDSA approach to dataspaces. 
-IDSA defines the architecture of dataspaces in the reference architecture model IDS RAM (current version 4). 
-One of the defined architectural components, the Vocabulary Hub, has the role of providing vocabularies to annotate and describe data assets and services. 
-These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
+IDSA defines the dataspace architecture in the IDS RAM [@IDSRAM2024].
+The Vocabulary Hub is one of the defined architectural components that provides vocabularies to annotate and describe data assets and services.
+These vocabularies form a common language to enable semantic interoperability among dataspace participants.
 The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions.
@@
 
-The Vocabulary Hub is one of the core components of any dataspace not only those designed in accordance with IDS RAM. 
-Other reference architecture models, such as GAIA-X architecture document 22.10 Release [@GaiaXArchitectureDocument2022] and Fraunhofer's Reference Architecture Model for the Industrial Data Space [@ReferenceArhictectureModel2017] also describe components with a similar role - the Federated Catalogue and Broker Service, respectively.
-Therefore, our approach designed around the IDS RAM is adaptable and applicable to data spaces designed in accordance with other reference architecture models.
+Other architecture reference models, such as the GAIA-X Architecture [@GaiaXArchitectureDocument2022] 
+describe components with a similar role: the Federated Catalogue and Broker Service.
+Therefore our approach  is adaptable and applicable to dataspaces designed in accordance with other reference architecture models as well.
 
 ## GraphDB Graph Database
-Ontotext GraphDB is a highly efficient, scalable and robust graph database with RDF and SPARQL support.
-It also offers data virtualization capabilities which enable direct access using SPARQL to relational databases, eliminating the need to replicate data.
-It can perform semantic inferencing at scale, which enables deriving new semantic facts from existing facts.
+[Ontotext GraphDB](https://graphdb.ontotext.com/) is a highly efficient, scalable and robust graph database with RDF and SPARQL support.
+It also offers data virtualization capabilities that enable direct SPARQL access to relational databases, eliminating the need to replicate data.
+It can perform semantic reasoning at scale and supports both incremental inference when triples are inserted, and retraction when triples are deleted.
+GraphDB handles massive semantic data loads, queries, and inferencing.
+Reasoning and query evaluation are performed over a persistent storage layer. 
+According to the [LDBC Semantic Publishing Benchmark](https://ldbcouncil.org/benchmarks/spb/), GraphDB is one of the most scalable OWL repositories currently available.
 
-GraphDB handles massive loads, queries, and inferencing in real time.
-Reasoning and query evaluation are performed over a persistent storage layer. Loading, reasoning, and query evaluation are executed extremely fast even against huge ontologies and knowledge bases.
-According to the [LDBC Semantic Publishing Benchmark](https://ldbcouncil.org/benchmarks/spb/), it is one of the most scalable OWL repositories currently available.
+In addition to numerous commercial projects, 
+GraphDB has been used for a variety of research projects, from astronomy to zoology [@alexievDiverseUsesSemantic2021].
 
 ## PoolParty Semantic Suite
-PoolParty Semantic Suite is a semantic middleware platform based on W3C standards, specifically the Semantic Web, which provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.
+[PoolParty Semantic Suite](PoolParty Semantic Suite) is a semantic middleware platform based on W3C Semantic Web standards 
+that provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.
 
-At the core of PoolParty is the Thesaurus Manager, which can be used to create and maintain RDF-based vocabularies, including ontologies, schemas and different kinds of knowledge models.
-These vocabularies can be applied for various use cases, from bridging data silos in enterprises by implementing data consolidation via the vocabularies, to annotating documents with semantic annotations to implement semantic search and recommendations which provide insights into data.
+At the core of PoolParty is the Thesaurus Manager that is used to create and maintain RDF-based vocabularies (thesauri and ontologies), 
+schemas and different kinds of knowledge models. 
+These vocabularies can be applied for various use cases, 
+from bridging data silos in enterprises by implementing data consolidation via the vocabularies, 
+to annotating documents with semantic annotations to implement semantic search and recommendations which provide insights into data.
 
@@
 ## The GraphDB Poolparty Integration
-The integration of GraphDB and PoolParty leverages synergies of a high-performance graph database with a platform for knowledge graph management.
+The integration of GraphDB and PoolParty leverages synergies of a high-performance graph database with a knowledge graph management platform.
 
-Both GraphDB and PoolParty are based on W3C recommendations like RDF, OWL and SPARQL.
+Both GraphDB and PoolParty are based on W3C recommendations (RDF, RDFS, OWL, SPARQL etc). 
 They therefore seamlessly integrate with each other, but also with other services and components in dataspaces which support these machine-readable standards for semantic interoperability.
@@
 ## Semantic Layer Approach
-The Vocabulary Hub, as IDS defines it in the IDS RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
-Those vocabularies are to be expressed in RDF and by syntactically compatible with OWL to allow for expressive semantics. 
-The IDS RAM does not require, but nonetheless allows for, extending the function of the Vocabulary hub towards providing ontology mappings [@FunctionalLayerIDSRAM2023b], which would enable dataspace connectors to automatically convert data between user-specific and standard data formats.
-Such automatic conversions are to be performed by Data Tranformation Apps by either the Data Consumers, the Data Providers, or both.
 
-We aim to achieve this through a Semantic Layer for dataspaces, 
-which provides vocabulary-based services for advanced and automated semantic descriptions of metadata and data. 
+The Vocabulary Hub is defined in the IDS RAM as a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies.
+Vocabularies are expressed in RDF and use RDFS and OWL for ontologies, and SKOS for thesauri.
+The IDS RAM does not require, but nonetheless allows extending the function of the Vocabulary Hub towards providing ontology mappings [@IDSARAMFunctional2023]
+that enable dataspace connectors to automatically convert data between user-specific and standard data formats.
+Such automatic conversions can be performed by Data Tranformation Apps by the Data Consumer, the Data Provider, or both.
+
+We aim to achieve this through a Semantic Layer for dataspaces
+that provides vocabulary-based services for advanced and automated semantic description of metadata and data.
 The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.
@@
 In the following, we discuss
-
 i) the services provided for the Vocabulary Hub component and
@@
   and to convert returned data to SPARQL result format.
-- The Data Consumer benefits from:
+
+The Data Consumer benefits from:
 - More powerful dataset discovery by using richer semantic queries, e.g.

### 2024-06-16 EBDVF-2024/paper.tex
edits

@@
 \author[1]{Robert David%
-%
+\corref{cor1}%
 }
@@
 \author[2]{Vladimir Alexiev%
-%
+\corref{cor1}%
 }
@@
 \end{abstract}
+
+\begin{highlights}
+\item The Vocabulary Hub is used in a dataspace to hold semantic
+vocabularies (ontologies and thesauri).\item Proposes and implements a
+Semantic Layers architecture that allows binding semantic vocabularies
+to incoming datasets.\item Uses semantic mapping languages (OBDA, R2RML,
+CSVW) to harmonize datasets dynamically.
+\end{highlights}
 
@@
 
-In this paper, we focus on the IDSA approach to dataspaces. IDSA defines
-the architecture of dataspaces in the reference architecture model IDS
-RAM (current version 4). One of the defined architectural components,
-the Vocabulary Hub, has the role of providing vocabularies to annotate
-and describe data assets and services. These vocabularies form a common
-language to enable semantic interoperability among the dataspace
+IDSA defines the dataspace architecture in the IDS RAM
+\citep{IDSRAM2024}. The Vocabulary Hub is one of the defined
+architectural components that provides vocabularies to annotate and
+describe data assets and services. These vocabularies form a common
+language to enable semantic interoperability among dataspace
 participants. The Vocabulary Hub is responsible for resolving
@@
 
-The Vocabulary Hub is one of the core components of any dataspace not
-only those designed in accordance with IDS RAM. Other reference
-architecture models, such as GAIA-X architecture document 22.10 Release
-\citep{GaiaXArchitectureDocument2022} and Fraunhofer's Reference
-Architecture Model for the Industrial Data Space
-\citep{ReferenceArhictectureModel2017} also describe components with a
-similar role - the Federated Catalogue and Broker Service, respectively.
-Therefore, our approach designed around the IDS RAM is adaptable and
-applicable to data spaces designed in accordance with other reference
-architecture models.
+Other architecture reference models, such as the GAIA-X Architecture
+\citep{GaiaXArchitectureDocument2022} describe components with a similar
+role: the Federated Catalogue and Broker Service. Therefore our approach
+is adaptable and applicable to dataspaces designed in accordance with
+other reference architecture models as well.
 
@@
 
-Ontotext GraphDB is a highly efficient, scalable and robust graph
-database with RDF and SPARQL support. It also offers data virtualization
-capabilities which enable direct access using SPARQL to relational
-databases, eliminating the need to replicate data. It can perform
-semantic inferencing at scale, which enables deriving new semantic facts
-from existing facts.
+\href{https://graphdb.ontotext.com/}{Ontotext GraphDB} is a highly
+efficient, scalable and robust graph database with RDF and SPARQL
+support. It also offers data virtualization capabilities that enable
+direct SPARQL access to relational databases, eliminating the need to
+replicate data. It can perform semantic reasoning at scale and supports
+both incremental inference when triples are inserted, and retraction
+when triples are deleted. GraphDB handles massive semantic data loads,
+queries, and inferencing. Reasoning and query evaluation are performed
+over a persistent storage layer. According to the
+\href{https://ldbcouncil.org/benchmarks/spb/}{LDBC Semantic Publishing
+Benchmark}, GraphDB is one of the most scalable OWL repositories
+currently available.
 
-GraphDB handles massive loads, queries, and inferencing in real time.
-Reasoning and query evaluation are performed over a persistent storage
-layer. Loading, reasoning, and query evaluation are executed extremely
-fast even against huge ontologies and knowledge bases. According to the
-\href{https://ldbcouncil.org/benchmarks/spb/}{LDBC Semantic Publishing
-Benchmark}, it is one of the most scalable OWL repositories currently
-available.
+In addition to numerous commercial projects, GraphDB has been used for a
+variety of research projects, from astronomy to zoology
+\citep{alexievDiverseUsesSemantic2021}.
 
@@
 
-PoolParty Semantic Suite is a semantic middleware platform based on W3C
-standards, specifically the Semantic Web, which provides a wide variety
-of functionality in the area of knowledge graph management, graph-based
-NLP, semantic search and recommender systems.
+\href{PoolParty\%20Semantic\%20Suite}{PoolParty Semantic Suite} is a
+semantic middleware platform based on W3C Semantic Web standards that
+provides a wide variety of functionality in the area of knowledge graph
+management, graph-based NLP, semantic search and recommender systems.
 
-At the core of PoolParty is the Thesaurus Manager, which can be used to
-create and maintain RDF-based vocabularies, including ontologies,
-schemas and different kinds of knowledge models. These vocabularies can
-be applied for various use cases, from bridging data silos in
-enterprises by implementing data consolidation via the vocabularies, to
-annotating documents with semantic annotations to implement semantic
-search and recommendations which provide insights into data.
+At the core of PoolParty is the Thesaurus Manager that is used to create
+and maintain RDF-based vocabularies (thesauri and ontologies), schemas
+and different kinds of knowledge models. These vocabularies can be
+applied for various use cases, from bridging data silos in enterprises
+by implementing data consolidation via the vocabularies, to annotating
+documents with semantic annotations to implement semantic search and
+recommendations which provide insights into data.
 
@@
 The integration of GraphDB and PoolParty leverages synergies of a
-high-performance graph database with a platform for knowledge graph
-management.
+high-performance graph database with a knowledge graph management
+platform.
 
-Both GraphDB and PoolParty are based on W3C recommendations like RDF,
-OWL and SPARQL. They therefore seamlessly integrate with each other, but
-also with other services and components in dataspaces which support
+Both GraphDB and PoolParty are based on W3C recommendations (RDF, RDFS,
+OWL, SPARQL etc). They therefore seamlessly integrate with each other,
+but also with other services and components in dataspaces which support
 these machine-readable standards for semantic interoperability.
@@
 
-The Vocabulary Hub, as IDS defines it in the IDS RAM, is a basic
-building block to achieve semantic interoperability in dataspaces by
-provisioning common vocabularies. Those vocabularies are to be expressed
-in RDF and by syntactically compatible with OWL to allow for expressive
-semantics. The IDS RAM does not require, but nonetheless allows for,
-extending the function of the Vocabulary hub towards providing ontology
-mappings \citep{FunctionalLayerIDSRAM2023b}, which would enable
-dataspace connectors to automatically convert data between user-specific
-and standard data formats. Such automatic conversions are to be
-performed by Data Tranformation Apps by either the Data Consumers, the
-Data Providers, or both.
+The Vocabulary Hub is defined in the IDS RAM as a basic building block
+to achieve semantic interoperability in dataspaces by provisioning
+common vocabularies. Vocabularies are expressed in RDF and use RDFS and
+OWL for ontologies, and SKOS for thesauri. The IDS RAM does not require,
+but nonetheless allows extending the function of the Vocabulary Hub
+towards providing ontology mappings \citep{IDSARAMFunctional2023} that
+enable dataspace connectors to automatically convert data between
+user-specific and standard data formats. Such automatic conversions can
+be performed by Data Tranformation Apps by the Data Consumer, the Data
+Provider, or both.
 
-We aim to achieve this through a Semantic Layer for dataspaces, which
+We aim to achieve this through a Semantic Layer for dataspaces that
 provides vocabulary-based services for advanced and automated semantic
-descriptions of metadata and data. The Semantic Layer is implemented as
+description of metadata and data. The Semantic Layer is implemented as
 services based on the GraphDB and PoolParty and provided as a one
@@
 
-In the following, we discuss
-
-\begin{enumerate}
-\def\labelenumi{\roman{enumi})}
-\tightlist
-\item
-  the services provided for the Vocabulary Hub component and
-\item
-  additional services as Vocabulary Hub extensions that implement a
-  Semantic Layer for dataspaces.
-\end{enumerate}
+In the following, we discuss i) the services provided for the Vocabulary
+Hub component and ii) additional services as Vocabulary Hub extensions
+that implement a Semantic Layer for dataspaces.
 
@@
   result format.
-\item
-  The Data Consumer benefits from:
+\end{itemize}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    More powerful dataset discovery by using richer semantic queries,
-    e.g.
-
-    \begin{itemize}
-    \tightlist
-    \item
-      ``Give me all time series related to compressor C123'' (equipment
-      instance)
-    \item
-      ``Give me all time series related to temperature'' (measured
-      quantity)
-    \item
-      ``Give me all time series of bearings'' (equipment kind or machine
-      part)
-    \end{itemize}
-  \item
-    Harmonized querying of datasets from multiple providers.
-  \end{itemize}
-\end{itemize}
+The Data Consumer benefits from: - More powerful dataset discovery by
+using richer semantic queries, e.g. - ``Give me all time series related
+to compressor C123'' (equipment instance) - ``Give me all time series
+related to temperature'' (measured quantity) - ``Give me all time series
+of bearings'' (equipment kind or machine part) - Harmonized querying of
+datasets from multiple providers.
 

### 2024-06-16 EBDVF-2024/paper.qmd
Add files via upload

@@
   The Vocabulary Hub service (part of the IDS RAM) provides a repository for ontologies and vocabularies.
-  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
+  We describe an approach of raising the role of the Vocabulary Hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
   and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
@@
 Other architecture reference models, such as the GAIA-X Architecture [@GaiaXArchitectureDocument2022] 
-describe components with a similar role: the Federated Catalogue and Broker Service.
+describe components with a similar role, namely the Federated Catalogue.
 Therefore our approach is adaptable and applicable to dataspaces designed in accordance with other reference architecture models as well.
@@
 ## PoolParty Semantic Suite
-[PoolParty Semantic Suite](PoolParty Semantic Suite) is a semantic middleware platform based on W3C Semantic Web standards 
+[PoolParty Semantic Suite](https://www.poolparty.biz/) is a semantic middleware platform based on W3C Semantic Web standards 
 that provides a wide variety of functionality in the area of knowledge graph management, graph-based NLP, semantic search and recommender systems.
@@
 In the following, we discuss
+
 i) the services provided for the Vocabulary Hub component and
@@
 The Data Consumer benefits from:
+
 - More powerful dataset discovery by using richer semantic queries, e.g.

### 2024-06-16 EBDVF-2024/paper.tex
Add files via upload

@@
 IDS RAM) provides a repository for ontologies and vocabularies. We
-describe an approach of raising the role of the vocabulary hub to also
+describe an approach of raising the role of the Vocabulary Hub to also
 allow richer metadata description (e.g.~the meaning of every column in a
@@
 \citep{GaiaXArchitectureDocument2022} describe components with a similar
-role: the Federated Catalogue and Broker Service. Therefore our approach
-is adaptable and applicable to dataspaces designed in accordance with
-other reference architecture models as well.
+role, namely the Federated Catalogue. Therefore our approach is
+adaptable and applicable to dataspaces designed in accordance with other
+reference architecture models as well.
 
@@
 
-\href{PoolParty\%20Semantic\%20Suite}{PoolParty Semantic Suite} is a
+\href{https://www.poolparty.biz/}{PoolParty Semantic Suite} is a
 semantic middleware platform based on W3C Semantic Web standards that
@@
 
-In the following, we discuss i) the services provided for the Vocabulary
-Hub component and ii) additional services as Vocabulary Hub extensions
-that implement a Semantic Layer for dataspaces.
+In the following, we discuss
+
+\begin{enumerate}
+\def\labelenumi{\roman{enumi})}
+\tightlist
+\item
+  the services provided for the Vocabulary Hub component and
+\item
+  additional services as Vocabulary Hub extensions that implement a
+  Semantic Layer for dataspaces.
+\end{enumerate}
 
@@
 
-The Data Consumer benefits from: - More powerful dataset discovery by
-using richer semantic queries, e.g. - ``Give me all time series related
-to compressor C123'' (equipment instance) - ``Give me all time series
-related to temperature'' (measured quantity) - ``Give me all time series
-of bearings'' (equipment kind or machine part) - Harmonized querying of
-datasets from multiple providers.
+The Data Consumer benefits from:
+
+\begin{itemize}
+\tightlist
+\item
+  More powerful dataset discovery by using richer semantic queries, e.g.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    ``Give me all time series related to compressor C123'' (equipment
+    instance)
+  \item
+    ``Give me all time series related to temperature'' (measured
+    quantity)
+  \item
+    ``Give me all time series of bearings'' (equipment kind or machine
+    part)
+  \end{itemize}
+\item
+  Harmonized querying of datasets from multiple providers.
+\end{itemize}
 

### 2024-08-16 EBDVF-2024/EBDVF-2024-presentation/presentation.qmd
Add files via upload

---
title: "Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces"
author: "Robert David, Petar Ivanov, Vladimir Alexiev"
date: 01-Oct-2024
subtitle: "3rd Workshop on Semantic Interoperability in Data Spaces"
format: revealjs
metadata-files:
  - ../_metadata.yml
---
# Acknowledgements
<div class="centered-text">
<img src="img/swc-logo-web.svg" width="45%">

<img src="img/Ontotext-Logo-vector.svg" width="45%">

<img src="img/underpin_logo_horizontal_icon_title_tagline_color_light_bg.svg" width="45%">

# Introduction

- the european data economy depends on the availability of data 
- technological foundation + services and tools to make use of it
- different industries and communities
- machine-learning systems need high-quality training data
    - predictive maintenance

Data space standards, like IDS, provide 
**technical foundation** and **data sovereignty** principles.

# Motivation

- interoperability is key for making use of shared data
- syntactic data exchange already solved, I.e. defined data formats and exchange standards
- semantic interoperability still an open challenge
- Starting point: IDS RAM and IM defines semantic metadata
- we need to extend it to cover **semantic interoperability for data**

# Approach
- Semantic Interoperability via a **Semantic Layer** for Dataspaces
    - based on IDS Vocabulary Hub
    - based on knowledge graph (KG) technologies
    - implemented by **GraphDB + PoolParty** products
- services + data for semantic interoperability
- data discovery & harmonisation
- semantic metadata + data for discovery and integration

# Use Cases
2 running projects:
DataBri-X
- energy community simulation and prediction
- legal knowledge graph + document analysis

UNDERPIN
- refinery predictive maintenance
- wind farms predictive maintenance

# Approach

Semantic Layer supports structured and unstructured data interlinked in the KG.
Structured
- Ontology-Based Data Access (OBDA)
- RDB to RDF Mapping Language (R2RML)
- CSV on the Web (CSVW)

Unstructured
- semantic document annotations via KG entities

# Approach

Semantic Layer supports inference services.

- reasoning based on W3C recommendations OWL & SHACL
- inference tagging expands semantic annotations in documents
- vocabulary crosswalks connect different similar vocabularies
- metadata inference to determine specific vocabularies for domains

# Integration Architecture
<div class="centered-text">
<img src="img/architecture.png" width="125%">

# Data Consumer Benefits

- harmonized and semantically integrated data from disparate providers
- richer metadata descriptions
- incoming data bound dynamically to semantic descriptions
- richer discoverability / easier data querying
- vocabulary crosswalks: expand queries and annotations by traversing the interlinked graph

# Summary and Future Work

A Semantic Layer enchances the Vocabulary Hub's role by interlinking structured and unstructured data from multiple providers and formats based on a harmonized model.
This improves the discoverability and comprehensibility of said data.

***Future work:***

- Implement use cases for richer discoverability, harmonized querying and support for different content types.
- Explore how ML can benefit regarding quality in practice when providing consolidated and cleaned data via dataspaces.
- Discuss how we can extend the IDS RAM with services to improve the support for semantic interoperability provided by our solution.

### 2024-10-01 EBDVF-2024/EBDVF-2024-presentation/presentation.qmd

@@
 metadata-files:
-  - ../_metadata.yml
+  - _metadata.yml
 ---
-# Acknowledgements
-<div class="centered-text">
-<img src="img/swc-logo-web.svg" width="45%">
-
-<img src="img/Ontotext-Logo-vector.svg" width="45%">
-
-<img src="img/underpin_logo_horizontal_icon_title_tagline_color_light_bg.svg" width="45%">
 
@@
 
-- the european data economy depends on the availability of data 
-- technological foundation + services and tools to make use of it
-- different industries and communities
-- machine-learning systems need high-quality training data
+- The European data economy depends on the availability of data 
+- Technological foundation + services and tools to make use of it
+- Different industries and communities
+- Machine-learning systems need high-quality training data
     - predictive maintenance
@@
 
-- interoperability is key for making use of shared data
-- syntactic data exchange already solved, I.e. defined data formats and exchange standards
-- semantic interoperability still an open challenge
+- Interoperability is key for making use of shared data
+- Syntactic data exchange already solved, I.e. defined data formats and exchange standards
+- Semantic interoperability still an open challenge
 - Starting point: IDS RAM and IM defines semantic metadata
@@
     - implemented by **GraphDB + PoolParty** products
-- services + data for semantic interoperability
+- Services + data for semantic interoperability
 - data discovery & harmonisation
-- semantic metadata + data for discovery and integration
+- Semantic metadata + data for discovery and integration
 
 # Use Cases
-2 running projects:
-DataBri-X
+Two running projects:
+
+**DataBri-X**
+
 - energy community simulation and prediction
@@
 
-UNDERPIN
+**UNDERPIN**
+
 - refinery predictive maintenance
@@
 Semantic Layer supports structured and unstructured data interlinked in the KG.
-Structured
-- Ontology-Based Data Access (OBDA)
-- RDB to RDF Mapping Language (R2RML)
+
+**Structured**
+
+::::{.columns}
+::: {.column width="40%"}
+- Ontology-Based Data Access
+- RDB to RDF Mapping Language
 - CSV on the Web (CSVW)
+:::
+::: {.column width="60%"}
+![](./img/skos.png){width=15%}
+![](./img/r2rml.png){width=15%}
+![](./img/csv.png){width=15%}
+:::
+::::
 
-Unstructured
-- semantic document annotations via KG entities
+**Unstructured**
+
+- Semantic document annotations via KG entities
 
@@
 
-- reasoning based on W3C recommendations OWL & SHACL
-- inference tagging expands semantic annotations in documents
-- vocabulary crosswalks connect different similar vocabularies
-- metadata inference to determine specific vocabularies for domains
+- Reasoning based on W3C recommendations OWL & SHACL
+- Inference tagging expands semantic annotations in documents
+- Vocabulary crosswalks connect different similar vocabularies
+- Metadata inference to determine specific vocabularies for domains
 
@@
 
-- harmonized and semantically integrated data from disparate providers
-- richer metadata descriptions
-- incoming data bound dynamically to semantic descriptions
-- richer discoverability / easier data querying
-- vocabulary crosswalks: expand queries and annotations by traversing the interlinked graph
+- Harmonized and semantically integrated data from disparate providers
+- Richer metadata descriptions
+- Incoming data bound dynamically to semantic descriptions
+- Richer discoverability / easier data querying
+- Vocabulary crosswalks: expand queries and annotations by traversing the interlinked graph
 
@@
 
-***Future work:***
+**Future work:**
 
